---
title: "Choosing Employment Weights: OEWS versus PUMS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing Employment Weights: OEWS versus PUMS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
helper_paths <- c("vignettes/_helpers.R", "_helpers.R")
source(helper_paths[file.exists(helper_paths)][[1]])

archive_303 <- system.file("extdata", "onet-mini", "db_30_3_text", package = "onet2r")
oews_path <- system.file("extdata", "oews-national-sample.csv", package = "onet2r")
```

Weights answer a population question. OEWS is usually the best choice when the
target is official occupation employment and wage context. PUMS is useful when
the target is a custom sample or demographic cell.

## Start with the Same O&#42;NET Score

```{r score}
abilities <- onet_archive_read(
  "30.3",
  "Abilities",
  path = archive_303,
  release_date = "2026-05-01"
)

score <- abilities |>
  filter(element_id == "1.A.1.a.1") |>
  transmute(onet_soc_code, measure_score = data_value)

score |>
  onet_kable()
```

## OEWS Weights

```{r oews-weights}
oews_weights <- onet_weight_panel_oews(
  onet_oews_national(2024, path = oews_path),
  year = 2024
)

oews_result <- onet_measure_aggregate(
  score,
  oews_weights,
  measure_id = "oral_comprehension_fixture"
)

oews_weights |>
  onet_kable()
oews_result |>
  select(-coverage, -provenance) |>
  onet_kable()
```

## PUMS Weights

```{r pums-weights}
pums <- tibble::tibble(
  SOCP = c("151252", "151252", "291141", "291141", "111011", "111011"),
  PWGTP = c(120, 80, 90, 110, 10, 30),
  sex = c("F", "M", "F", "M", "F", "M")
)

pums_weights <- onet_weight_panel_pums(
  pums,
  year = 2022,
  group = "sex"
)

pums_f <- onet_measure_aggregate(
  score,
  pums_weights,
  measure_id = "oral_comprehension_fixture",
  cell = list(sex = "F")
)
pums_m <- onet_measure_aggregate(
  score,
  pums_weights,
  measure_id = "oral_comprehension_fixture",
  cell = list(sex = "M")
)

pums_weights |>
  onet_kable()
```

## Compare the Answers

```{r compare}
comparison <- tibble::tibble(
  population = c("OEWS national", "PUMS-style F", "PUMS-style M"),
  aggregate = c(oews_result$aggregate, pums_f$aggregate, pums_m$aggregate),
  coverage = c(
    oews_result$employment_coverage_share,
    pums_f$employment_coverage_share,
    pums_m$employment_coverage_share
  )
)

comparison |>
  onet_kable()
```

```{r weights-chart, fig.alt = "Horizontal dot chart comparing OEWS and PUMS-style weighted aggregates."}
ggplot2::ggplot(comparison, ggplot2::aes(
  x = aggregate,
  y = stats::reorder(population, aggregate)
)) +
  ggplot2::geom_segment(
    ggplot2::aes(x = 0, xend = aggregate, yend = population),
    color = onet2r_colors[["light_gray"]],
    linewidth = 1.1
  ) +
  ggplot2::geom_point(color = onet2r_colors[["teal"]], size = 3.5) +
  ggplot2::labs(
    title = "Weight Source Changes the Population",
    subtitle = "The score is fixed; the target population changes.",
    x = "Aggregate score",
    y = NULL
  ) +
  onet2r_theme()
```

If your claim is about the national labor market, OEWS is the natural default.
If your claim is about a subgroup, geography, or survey-defined population,
build a PUMS weight panel and state the cell explicitly.
