---
title: "From Tasks to Occupations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From Tasks to Occupations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
helper_paths <- c(
  system.file("vignette-helpers.R", package = "onet2r"),
  "inst/vignette-helpers.R",
  "../inst/vignette-helpers.R",
  "vignette-helpers.R"
)
helper_path <- helper_paths[nzchar(helper_paths) & file.exists(helper_paths)]
if (length(helper_path) == 0) {
  stop("Could not find vignette-helpers.R.", call. = FALSE)
}
source(helper_path[[1]])

archive_dir <- onet2r_inst_path("extdata", "onet-mini", "db_30_3_text")
```

Many user measures start at the task level: a researcher labels tasks, a model
scores task text, or a team codes task exposure manually. Occupation-level
analysis needs one score per occupation. `onet_task_to_occupation()` performs
that mechanical rollup using O&#42;NET task ratings.

## Read Task Statements and Ratings

```{r tasks}
tasks <- onet_archive_read(
  "30.3",
  "Task Statements",
  path = archive_dir,
  release_date = "2026-05-01"
)
ratings <- onet_archive_read(
  "30.3",
  "Task Ratings",
  path = archive_dir,
  release_date = "2026-05-01"
)

tasks |>
  select(onet_soc_code, task_id, task_type, task) |>
  onet_kable()

ratings |>
  select(onet_soc_code, task_id, scale_id, scale_name, data_value) |>
  head(8) |>
  onet_kable()
```

## Validate Task Scores

```{r measure}
task_scores <- tibble::tibble(
  task_id = c("1001", "1002", "2001"),
  score = c(0.80, 0.40, 0.20)
)

measure <- onet_measure(
  task_scores,
  key = "task_id",
  score = "score",
  key_type = "task",
  universe = tasks$task_id,
  measure_id = "stylized_task_score",
  release_version = "30.3"
)

onet_coverage(measure) |>
  onet_kable()
```

## Roll Up with Relevance Weights

```{r rollup-relevance}
core_only <- onet_task_to_occupation(
  measure,
  task_ratings = ratings,
  task_metadata = tasks,
  weight_scale = "RT",
  include_supplemental = FALSE
)

core_plus_supplemental <- onet_task_to_occupation(
  measure,
  task_ratings = ratings,
  task_metadata = tasks,
  weight_scale = "RT",
  include_supplemental = TRUE
)

core_only |>
  select(onet_soc_code, soc_code, n_tasks, total_task_weight, measure_score) |>
  onet_kable()
core_plus_supplemental |>
  select(onet_soc_code, soc_code, n_tasks, total_task_weight, measure_score) |>
  onet_kable()
```

## Interpret the Plumbing Choice

```{r compare-task-handling}
comparison <- bind_rows(
  core_only |> mutate(rule = "Core tasks only"),
  core_plus_supplemental |> mutate(rule = "Core plus Supplemental")
) |>
  select(rule, onet_soc_code, n_tasks, total_task_weight, measure_score)

comparison |>
  onet_kable()
```

```{r task-chart, fig.alt = "Grouped dot chart of occupation scores under core-only and core-plus-supplemental task rules."}
ggplot2::ggplot(comparison, ggplot2::aes(
  x = measure_score,
  y = onet_soc_code,
  color = rule
)) +
  ggplot2::geom_point(
    ggplot2::aes(shape = rule),
    size = 3,
    position = ggplot2::position_dodge(width = 0.45)
  ) +
  ggplot2::labs(
    title = "Task Handling Can Change Occupation Scores",
    subtitle = "The task score is fixed; the rollup rule changes.",
    x = "Occupation score",
    y = "O*NET-SOC code",
    color = NULL,
    shape = NULL
  ) +
  ggplot2::scale_color_manual(
    values = c(
      "Core tasks only" = onet2r_colors[["teal"]],
      "Core plus Supplemental" = onet2r_colors[["amber"]]
    )
  ) +
  onet2r_theme()
```

The task score did not change; only the rollup rule did. A clear write-up keeps
the scoring choice separate from the package mechanics.
