---
title: "Getting Started with onet2r"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with onet2r}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
options(
  cli.num_colors = 1,
  crayon.enabled = FALSE,
  pillar.bold = FALSE,
  pillar.subtle = FALSE
)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
helper_paths <- c("vignettes/_helpers.R", "_helpers.R")
source(helper_paths[file.exists(helper_paths)][[1]])

has_onet_key <- nzchar(Sys.getenv("ONET_API_KEY"))
run_live <- has_onet_key && identical(Sys.getenv("ONET2R_RUN_LIVE_VIGNETTES"), "true")

sample_oews <- system.file(
  "extdata",
  "oews-national-sample.csv",
  package = "onet2r"
)

archive_base <- system.file("extdata", "onet-mini", package = "onet2r")
archive_302 <- file.path(archive_base, "db_30_2_text")
archive_303 <- file.path(archive_base, "db_30_3_text")
```

`onet2r` helps you move from occupation search to analysis-ready O&#42;NET tables,
archived O&#42;NET releases, and BLS OEWS wage and employment context. Current
O&#42;NET Web Services calls require a free API key from the
[O&#42;NET developer portal](https://services.onetcenter.org/developer/). Store it
in `.Renviron` as `ONET_API_KEY=your-api-key-here` so scripts do not contain
secrets.

The live Web Services examples are opt-in for package builds because CRAN and CI
should not depend on an external API. Everything else in this article uses local
files through actual `onet2r` functions.

```{r auth-status}
tibble::tibble(
  setting = c("ONET_API_KEY configured", "live vignette API calls enabled"),
  value = c(has_onet_key, run_live)
) |>
  onet_kable()
```

## Search the Current O&#42;NET API

When live calls are enabled, the first step is usually `onet_search()`, followed
by a detail endpoint such as `onet_skills()` or `onet_abilities()`.

```{r live-api-search}
if (run_live) {
  onet_search("software developer", start = 1, end = 5) |>
    onet_kable()
} else {
  tibble::tibble(
    live_api_example = "skipped",
    reason = "Set ONET_API_KEY and ONET2R_RUN_LIVE_VIGNETTES=true to run onet_search()."
  ) |>
    onet_kable()
}
```

```{r live-api-details}
if (run_live) {
  onet_skills("15-1252.00", start = 1, end = 5) |>
    onet_kable()
} else {
  tibble::tibble(
    live_api_example = "skipped",
    reason = "Set ONET_API_KEY and ONET2R_RUN_LIVE_VIGNETTES=true to run onet_skills()."
  ) |>
    onet_kable()
}
```

## Read an Archived O&#42;NET Table

The Web Services API serves the current release. Historical work uses the
downloadable archive tables. `onet_archive_read()` normalizes those archive files
into tibbles with stable columns.

```{r archive-read}
abilities <- onet_archive_read(
  "30.3",
  "Abilities",
  path = archive_303,
  release_date = "2026-05-01"
)

abilities |>
  select(
    release_version,
    onet_soc_code,
    soc_code,
    element_id,
    element_name,
    data_value,
    source_date,
    domain_source
  ) |>
  head(8) |>
  onet_kable()
```

## Add Labor-Market Context

BLS OEWS estimates add employment and wage scale to O&#42;NET occupation rows.
The modern weighting path creates an explicit reference-SOC weight panel.

```{r oews-weights}
oews <- onet_oews_national(2024, path = sample_oews)
weights <- onet_weight_panel_oews(oews, year = 2024)

weights |>
  onet_kable()
```

To answer a concrete question, take one ability, treat the O&#42;NET values as the
user-supplied occupation score, and aggregate with OEWS employment.

```{r aggregate-ability}
oral_scores <- abilities |>
  filter(element_id == "1.A.1.a.1") |>
  transmute(onet_soc_code, measure_score = data_value)

oral_aggregate <- onet_measure_aggregate(
  oral_scores,
  weights,
  measure_id = "oral_comprehension_fixture"
)

oral_aggregate |>
  select(-coverage, -provenance) |>
  onet_kable()

onet_provenance(oral_aggregate) |>
  onet_kable()
```

## Compare Two Archive Releases

For historical analysis, build a panel, reconcile adjacent releases, and inspect
comparability flags before interpreting changes.

```{r longitudinal-preview}
panel <- onet_panel(
  "Abilities",
  versions = c("30.2", "30.3"),
  scale = "IM",
  archives = c(`30.2` = archive_302, `30.3` = archive_303),
  release_dates = c(`30.2` = "2026-02-01", `30.3` = "2026-05-01")
)

changes <- onet_panel_reconcile(
  panel,
  bridge = onet_crosswalk_bridge("2019", "2019")
)

changes |>
  select(
    to_soc_code,
    element_name,
    from_value,
    to_value,
    value_change,
    change_type,
    safely_comparable
  ) |>
  arrange(desc(abs(value_change))) |>
  head(8) |>
  onet_kable()
```

```{r getting-started-chart, fig.alt = "Horizontal bar chart of change classifications in the getting started example."}
plot_data <- changes |>
  count(change_type, name = "rows") |>
  arrange(desc(rows))

ggplot2::ggplot(plot_data, ggplot2::aes(
  x = stats::reorder(change_type, rows),
  y = rows,
  fill = change_type
)) +
  ggplot2::geom_col(width = 0.65, show.legend = FALSE) +
  ggplot2::coord_flip() +
  onet2r_discrete_fill() +
  ggplot2::labs(
    title = "What Kind of Change Was Observed?",
    subtitle = "Interpret the classification before the numeric change.",
    x = NULL,
    y = "Rows"
  ) +
  onet2r_theme()
```

## Next Steps

- Read `vignette("longitudinal-onet-background", package = "onet2r")` before
  interpreting cross-release changes.
- Read `vignette("longitudinal-archives", package = "onet2r")` to assemble and
  reconcile archive panels.
- Read `vignette("oews-wage-context", package = "onet2r")` to choose and apply
  employment weights.
- Read `vignette("stress-testing-exposure-measure", package = "onet2r")` to
  stress test a user-supplied measure.
