---
title: "Getting Started with onet2r"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with onet2r}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(onet2r)
library(dplyr)
library(tibble)

has_onet_key <- nzchar(Sys.getenv("ONET_API_KEY"))
run_live <- has_onet_key && identical(Sys.getenv("ONET2R_RUN_LIVE_VIGNETTES"), "true")
```

`onet2r` helps you move from occupation search to analysis-ready O*NET
tables. The Web Services API requires a free API key from the
[O*NET developer portal](https://services.onetcenter.org/developer/). Store it
in `.Renviron` as `ONET_API_KEY=your-api-key-here` so you do not put secrets in
scripts.

This article uses small representative tibbles so the pkgdown site always shows
output. To run the same chunks against the live API while building locally, set
both `ONET_API_KEY` and `ONET2R_RUN_LIVE_VIGNETTES=true`.

```{r auth-status}
tibble(
  setting = c("ONET_API_KEY configured", "live vignette calls enabled"),
  value = c(has_onet_key, run_live)
)
```

## Search for Occupations

Search is usually the first step. Results include an O*NET-SOC code and an
occupation title.

```{r search}
software_jobs <- if (run_live) {
  onet_search("software developer", start = 1, end = 5)
} else {
  tibble(
    code = c("15-1252.00", "15-1251.00", "15-2051.00"),
    title = c(
      "Software Developers",
      "Computer Programmers",
      "Data Scientists"
    )
  )
}

head(software_jobs, 5)
```

## Retrieve Occupation Details

Most detail endpoints return tibbles. That makes them easy to inspect, join, and
summarize.

```{r details}
skills <- if (run_live) {
  onet_skills("15-1252.00", start = 1, end = 5)
} else {
  tibble(
    id = c("2.A.1.a", "2.A.1.b", "2.B.3.m"),
    name = c(
      "Reading Comprehension",
      "Active Listening",
      "Programming"
    ),
    description = c(
      "Understanding written sentences and paragraphs in work-related documents.",
      "Giving full attention to what other people are saying.",
      "Writing computer programs for various purposes."
    ),
    value = c(4.12, 4.00, 4.38),
    scale = c("Importance", "Importance", "Importance")
  )
}

head(skills, 5)
```

The same pattern works for knowledge, abilities, tasks, work activities, work
styles, work context, interests, technology, education, and job zones.

## Work with O*NET Tables

The O*NET database tables expose lower-level source data. Use
`onet_tables()` to find available tables and `onet_table_info()` to inspect
columns before downloading a full table.

```{r tables}
tables <- if (run_live) {
  onet_tables()
} else {
  tibble(
    id = c("Occupation Data", "Abilities", "Skills", "Task Ratings"),
    title = c("Occupation Data", "Abilities", "Skills", "Task Ratings")
  )
}

head(tables, 6)
```

```{r table-info}
table_info <- if (run_live) {
  onet_table_info("Abilities")
} else {
  tibble(
    name = c("onetsoc_code", "element_id", "scale_id", "data_value"),
    title = c("O*NET-SOC Code", "Element ID", "Scale ID", "Data Value"),
    type = c("character", "character", "character", "decimal"),
    format = c(NA, NA, NA, "n.n")
  )
}

head(table_info, 8)
```

## Join Labor-Market Context

O*NET describes what occupations do. BLS OEWS adds employment and wage scale.
This example uses the package's bundled OEWS sample so it always runs.

```{r oews}
sample_oews <- system.file(
  "extdata",
  "oews-national-sample.csv",
  package = "onet2r"
)

oews <- onet_oews_national(2024, path = sample_oews)

occupations <- tibble(
  code = c("15-1252.00", "29-1141.00", "11-1011.00"),
  title = c("Software Developers", "Registered Nurses", "Chief Executives")
)

occupation_wages <- onet_join_oews(occupations, oews = oews)
head(occupation_wages, 5)
```

## Summarize Weighted O*NET Records

Employment-weighted summaries are useful when the same descriptor appears
across occupations with very different labor-market sizes.

```{r weighted-summary}
skill_rows <- tibble(
  code = c("15-1252.00", "15-1252.00", "29-1141.00", "29-1141.00"),
  element_id = c("2.A.1.a", "2.A.1.b", "2.A.1.a", "2.A.1.b"),
  element_name = c(
    "Reading Comprehension",
    "Active Listening",
    "Reading Comprehension",
    "Active Listening"
  ),
  data_value = c(4.12, 4.00, 3.88, 4.25)
)

weighted_skills <- onet_weighted_summary(
  skill_rows,
  group = c("element_id", "element_name"),
  value = "data_value",
  oews = oews
)

head(weighted_skills, 5)
```

## Build Longitudinal Archive Comparisons

The API serves the current release. To compare releases over time, use the
archive helpers. See the longitudinal articles for the methodological cautions.

```{r longitudinal-preview}
panel <- tibble(
  release_version = rep(c("1.0", "2.0"), each = 2),
  release_date = rep(as.Date(c("2020-01-01", "2021-01-01")), each = 2),
  soc_vintage = "2019",
  domain = "Abilities",
  onet_soc_code = rep(c("15-1252.00", "29-1141.00"), 2),
  soc_code = rep(c("15-1252", "29-1141"), 2),
  title = rep(c("Software Developers", "Registered Nurses"), 2),
  element_id = "1.A.1.a.1",
  element_name = "Oral Comprehension",
  scale_id = "IM",
  data_value = c(4.0, 4.2, 4.4, 4.2),
  source_date = as.Date(c("2020-01-01", "2020-01-01", "2021-01-01", "2020-01-01")),
  domain_source = "Analyst"
)

bridge <- tibble(
  from_vintage = "2019",
  to_vintage = "2019",
  from_soc_code = c("15-1252", "29-1141"),
  to_soc_code = c("15-1252", "29-1141"),
  from_title = c("Software Developers", "Registered Nurses"),
  to_title = c("Software Developers", "Registered Nurses"),
  step_count = 0L,
  map_type = "one_to_one",
  crosswalk_weight = 1
)

changes <- onet_panel_reconcile(panel, bridge)
head(changes, 5)
```

## Next Steps

- Read `vignette("longitudinal-onet-background", package = "onet2r")` before
  interpreting cross-release changes.
- Read `vignette("longitudinal-archives", package = "onet2r")` to assemble and
  reconcile archive panels.
- Read `vignette("oews-wage-context", package = "onet2r")` to add BLS wage and
  employment estimates.
