---
title: "Getting Started with onet2r"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with onet2r}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(onet2r)
library(dplyr)

has_onet_key <- nzchar(Sys.getenv("ONET_API_KEY"))
run_live <- has_onet_key && identical(Sys.getenv("ONET2R_RUN_LIVE_VIGNETTES"), "true")

sample_oews <- system.file(
  "extdata",
  "oews-national-sample.csv",
  package = "onet2r"
)

archive_302 <- system.file(
  "extdata",
  "onet-mini",
  "db_30_2_text",
  package = "onet2r"
)

archive_303 <- system.file(
  "extdata",
  "onet-mini",
  "db_30_3_text",
  package = "onet2r"
)
```

`onet2r` helps you move from occupation search to analysis-ready O&#42;NET tables,
archived O&#42;NET releases, and BLS OEWS wage and employment context. Current
O&#42;NET Web Services calls require a free API key from the
[O&#42;NET developer portal](https://services.onetcenter.org/developer/). Store it
in `.Renviron` as `ONET_API_KEY=your-api-key-here` so you do not put secrets in
scripts.

The live Web Services examples are opt-in for package builds, because CRAN and
most CI environments should not depend on an external API. The rest of this
article uses local files through actual `onet2r` functions, so every displayed
table is computed when the vignette is knitted.

```{r auth-status}
tibble::tibble(
  setting = c("ONET_API_KEY configured", "live vignette API calls enabled"),
  value = c(has_onet_key, run_live)
)
```

## Search the Current O&#42;NET API

When live calls are enabled, the first step is usually `onet_search()`, followed
by a detail endpoint such as `onet_skills()` or `onet_abilities()`.

```{r live-api-search}
if (run_live) {
  onet_search("software developer", start = 1, end = 5)
} else {
  tibble::tibble(
    live_api_example = "skipped",
    reason = "Set ONET_API_KEY and ONET2R_RUN_LIVE_VIGNETTES=true to run onet_search()."
  )
}
```

```{r live-api-details}
if (run_live) {
  onet_skills("15-1252.00", start = 1, end = 5)
} else {
  tibble::tibble(
    live_api_example = "skipped",
    reason = "Set ONET_API_KEY and ONET2R_RUN_LIVE_VIGNETTES=true to run onet_skills()."
  )
}
```

## Read an Archived O&#42;NET Table

The Web Services API serves the current release. Historical work uses the
downloadable archive tables. `onet_archive_read()` normalizes those archive files
into tibbles with stable columns.

```{r archive-read}
abilities <- onet_archive_read(
  "30.3",
  "Abilities",
  path = archive_303,
  release_date = "2026-05-01"
)

abilities |>
  select(
    release_version,
    onet_soc_code,
    soc_code,
    element_id,
    element_name,
    data_value,
    source_date,
    domain_source
  ) |>
  head(8) |>
  print(width = Inf)
```

## Add Labor-Market Context

BLS OEWS estimates add employment and wage scale to O&#42;NET occupation rows.
`onet_oews_national()` reads the OEWS file, and `onet_join_oews()` joins by a
normalized SOC code.

```{r oews-join}
oews <- onet_oews_national(2024, path = sample_oews)

occupations <- abilities |>
  distinct(code = onet_soc_code, title)

occupation_wages <- onet_join_oews(occupations, oews = oews)

occupation_wages |>
  select(code, title, soc_code, tot_emp, a_median)
```

## Summarize O&#42;NET Rows with Employment Weights

`onet_weighted_summary()` lets you summarize descriptor rows by employment or
wage context. Here it computes employment-weighted ability means from actual
archive rows and the bundled OEWS sample.

```{r weighted-summary}
onet_weighted_summary(
  abilities,
  group = c("element_id", "element_name"),
  value = "data_value",
  occupation_code = "onet_soc_code",
  oews = oews
) |>
  print(width = Inf)
```

## Compare Two Archive Releases

For historical analysis, build a panel, reconcile adjacent releases, and inspect
comparability flags before interpreting changes.

```{r longitudinal-preview}
panel <- onet_panel(
  "Abilities",
  versions = c("30.2", "30.3"),
  scale = "IM",
  archives = c(`30.2` = archive_302, `30.3` = archive_303),
  release_dates = c(`30.2` = "2026-02-01", `30.3` = "2026-05-01")
)

changes <- onet_panel_reconcile(
  panel,
  bridge = onet_crosswalk_bridge("2019", "2019")
)

changes |>
  select(
    to_soc_code,
    element_name,
    from_value,
    to_value,
    value_change,
    change_type,
    safely_comparable
  ) |>
  arrange(desc(abs(value_change))) |>
  head(8) |>
  print(width = Inf)
```

## Next Steps

- Read `vignette("longitudinal-onet-background", package = "onet2r")` before
  interpreting cross-release changes.
- Read `vignette("longitudinal-archives", package = "onet2r")` to assemble and
  reconcile archive panels.
- Read `vignette("oews-wage-context", package = "onet2r")` to add BLS wage and
  employment estimates.
