---
title: "How to Fool Yourself with O*NET over Time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Fool Yourself with O*NET over Time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
helper_paths <- c(
  "inst/vignette-helpers.R",
  "vignette-helpers.R",
  system.file("vignette-helpers.R", package = "onet2r")
)
helper_path <- helper_paths[nzchar(helper_paths) & file.exists(helper_paths)][[1]]
source(helper_path)

archive_base <- system.file("extdata", "onet-mini", package = "onet2r")
```

A common mistake is to treat every release-to-release difference as a fresh
within-occupation change. O&#42;NET archives contain source dates, domain sources,
suppression flags, and taxonomy vintages. Ignoring those fields can turn a
carryforward, transition row, or crosswalk seam into a misleading trend.

<div class="onet-callout onet-callout-warning">
<strong>Do not start with the coefficient.</strong>
Start with the audit fields. If a result depends on rows that were carried
forward, bridged across a split, or marked as transition data, the caveat
belongs next to the estimate.
</div>

## The Naive Difference

```{r naive-panel}
panel <- onet_panel(
  "Abilities",
  versions = c("30.2", "30.3"),
  scale = "IM",
  archives = c(
    `30.2` = file.path(archive_base, "db_30_2_text"),
    `30.3` = file.path(archive_base, "db_30_3_text")
  ),
  release_dates = c(`30.2` = "2026-02-01", `30.3` = "2026-05-01")
)

naive <- panel |>
  select(release_version, onet_soc_code, element_id, element_name, data_value)

naive <- naive |>
  filter(release_version == "30.2") |>
  select(onet_soc_code, element_id, element_name, from_value = data_value) |>
  inner_join(
    naive |>
      filter(release_version == "30.3") |>
      select(onet_soc_code, element_id, to_value = data_value),
    by = join_by(onet_soc_code, element_id),
    relationship = "one-to-one"
  ) |>
  mutate(naive_change = to_value - from_value) |>
  arrange(desc(abs(naive_change)))

naive |>
  head(8) |>
  onet_kable()
```

The naive table is useful as a screening tool. It is not enough for an
interpretation.

## The Reconciled Difference

```{r reconciled-difference}
changes <- onet_panel_reconcile(panel, onet_crosswalk_bridge("2019", "2019"))

changes |>
  select(
    to_soc_code,
    element_name,
    value_change,
    change_type,
    from_source_date,
    to_source_date,
    method_break,
    safely_comparable
  ) |>
  arrange(desc(abs(value_change))) |>
  head(8) |>
  onet_kable()
```

Now the same differences have labels. A change without a source-date change is
not as strong as a change with fresh source data. A method break should be
treated as a warning rather than as clean within-occupation change.

<div class="onet-callout onet-callout-tip">
<strong>A useful rule of thumb.</strong>
Use `value_change` to find rows worth inspecting, then use `change_type`,
`method_break`, and `safely_comparable` to decide what those rows can support.
</div>

## The Taxonomy Seam

```{r taxonomy-seam}
cross_panel <- onet_panel(
  "Abilities",
  versions = c("24.3", "25.1"),
  scale = "IM",
  archives = c(
    `24.3` = file.path(archive_base, "db_24_3_text"),
    `25.1` = file.path(archive_base, "db_25_1_text")
  ),
  release_dates = c(`24.3` = "2020-08-01", `25.1` = "2020-11-01")
)

bridge <- tibble::tibble(
  from_vintage = "2010",
  to_vintage = "2019",
  from_onet_soc_code = c("15-1132.00", "15-1132.00", "29-1141.00"),
  to_onet_soc_code = c("15-1252.00", "15-1253.00", "29-1141.00"),
  map_type = c("split", "split", "one_to_one"),
  crosswalk_weight = c(0.5, 0.5, 1)
)

seam <- onet_panel_reconcile(cross_panel, bridge)

seam |>
  select(
    from_onet_soc_code,
    to_onet_soc_code,
    element_name,
    change_type,
    transition_data,
    crosswalk_uncertain,
    safely_comparable
  ) |>
  onet_kable()
```

```{r fool-yourself-chart, fig.alt = "Horizontal bar chart comparing safe and unsafe comparisons at a taxonomy seam."}
safe_counts <- seam |>
  mutate(comparability = if_else(safely_comparable, "Safe", "Not safe")) |>
  count(comparability, name = "rows")

ggplot2::ggplot(safe_counts, ggplot2::aes(
  x = comparability,
  y = rows,
  fill = comparability
)) +
  ggplot2::geom_col(width = 0.6, show.legend = FALSE) +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(
    values = c("Safe" = onet2r_colors[["teal"]], "Not safe" = onet2r_colors[["rose"]])
  ) +
  ggplot2::labs(
    title = "Taxonomy Seams Reduce Safe Comparisons",
    subtitle = "A split occupation carries extra uncertainty even when codes can be bridged.",
    x = NULL,
    y = "Rows"
  ) +
  onet2r_theme()
```

Before making a historical claim, ask: Did the value change? Did the source date
change? Did the occupation cross a taxonomy seam? Was either row transition data
or suppressed? The package gives you those fields so the caveats do not get lost.
