---
title: "Importing Published Exposure Scores"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Importing Published Exposure Scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
options(
  cli.num_colors = 1,
  crayon.enabled = FALSE,
  pillar.bold = FALSE,
  pillar.subtle = FALSE
)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
helper_paths <- c(
  system.file("vignette-helpers.R", package = "onet2r"),
  "inst/vignette-helpers.R",
  "../inst/vignette-helpers.R",
  "vignette-helpers.R"
)
helper_path <- helper_paths[nzchar(helper_paths) & file.exists(helper_paths)]
if (length(helper_path) == 0) {
  stop("Could not find vignette-helpers.R.", call. = FALSE)
}
source(helper_path[[1]])

archive_303 <- onet2r_inst_path("extdata", "onet-mini", "db_30_3_text")
oews_path <- onet2r_inst_path("extdata", "oews-national-sample.csv")
```

`onet2r` does not ship or endorse an exposure measure. When you want to work
with a published one, two adapters read the authors' occupation-level files and
turn them into the same task-grain `onet_measure()` object that a hand-built
score would produce:

- `onet_import_eloundou()` reads the GPT exposure table from Eloundou et al.
  [-@eloundou2023gpts], keyed on 8-digit O&#42;NET-SOC codes.
- `onet_import_felten_aioe()` reads the AI Occupational Exposure (AIOE) scores
  from Felten, Raj, and Seamans [-@felten2021aioe], keyed on 6-digit SOC codes.

Both are thin adapters. They select a score column, standardize the occupation
code, join it to the tasks of a panel you supply, and record where the file came
from. They do not rescale, average, or otherwise transform the published values,
and the package never bundles the files.

<div class="onet-callout onet-callout-warning">
<strong>Stylized values.</strong>
To keep this article offline, it writes small local extracts with the published
column names and made-up values. None of the numbers below are the published
exposure scores.
</div>

## Choose the Task Panel

The panel sets the grain of the result: every distinct occupation and task pair
in it receives a score. Here it comes from the relevance ratings of a pinned
release in the bundled fixtures.

```{r task-panel}
tasks <- onet_archive_read(
  "30.3",
  "Task Statements",
  path = archive_303,
  release_date = "2026-05-01"
)
ratings <- onet_archive_read(
  "30.3",
  "Task Ratings",
  path = archive_303,
  release_date = "2026-05-01"
)

task_panel <- distinct(ratings, onet_soc_code, task_id)

task_panel |>
  knitr::kable(align = "l")
```

## Import an Occupation-Level Score

```{r stylized-extract}
eloundou_extract <- tempfile(fileext = ".csv")
utils::write.csv(
  data.frame(
    `O*NET-SOC Code` = c("15-1252.00", "29-1141.00", "11-1011.00"),
    human_rating_beta = c(0.60, 0.20, 0.40),
    dv_rating_beta = c(0.70, 0.30, 0.50),
    check.names = FALSE
  ),
  eloundou_extract,
  row.names = FALSE
)

exposure <- onet_import_eloundou(
  task_panel,
  path = eloundou_extract,
  score = "human_rating_beta",
  as_of = "stylized extract for this article",
  release_version = "30.3"
)

exposure$data |>
  select(onet_soc_code, task_id, measure_key, measure_score) |>
  knitr::kable(digits = 3, align = "l")
```

Every task inherits its occupation's published score. Both Software Developers
tasks get the same value, because the source is occupation-level. This is the
structurally blind construction that task-aware measures are usually contrasted
against, so treat it as a benchmark rather than a task-level estimate. The
published file offers several definitions, such as `human_rating_beta` and the
model-labeled `dv_rating_beta`; `score` selects one, and choosing it is your
decision.

## Check the Source Receipt

Each imported measure records a receipt with the file's SHA-256 digest, size,
retrieval time, and any `as_of` label.

```{r receipt}
receipt <- exposure$metadata$source_receipt

tibble::tibble(
  field = c("provenance_status", "as_of", "file_size", "actual_sha256"),
  value = c(
    receipt$provenance_status,
    receipt$as_of,
    format(receipt$file_size),
    receipt$actual_sha256
  )
) |>
  knitr::kable(align = "l")
```

Record the digest the first time you read a file, and pass it back as
`expected_sha256` in later runs. A changed file is rejected before it is parsed.

```{r digest-check}
recorded_digest <- receipt$actual_sha256

verified <- onet_import_eloundou(
  task_panel,
  path = eloundou_extract,
  expected_sha256 = recorded_digest,
  release_version = "30.3"
)
rejected <- tryCatch(
  onet_import_eloundou(
    task_panel,
    path = eloundou_extract,
    expected_sha256 = strrep("0", 64)
  ),
  error = conditionMessage
)

tibble::tibble(
  check = c("Recorded digest accepted", "Wrong digest rejected"),
  result = c(
    inherits(verified, "onet_measure"),
    grepl("digest mismatch", rejected, fixed = TRUE)
  )
) |>
  knitr::kable(align = "l")
```

## Roll Up and Weight

The imported object is an ordinary task-grain measure, so the usual plumbing
applies.

```{r rollup}
occupation_exposure <- onet_task_to_occupation(
  exposure,
  task_ratings = ratings,
  task_metadata = tasks
)

occupation_exposure |>
  select(onet_soc_code, n_tasks, total_task_weight, measure_score) |>
  knitr::kable(digits = 3, align = "l")

weights <- onet_weight_panel_oews(
  onet_oews_national(2024, path = oews_path),
  year = 2024
)

national <- onet_measure_aggregate(occupation_exposure, weights)

national |>
  select(-coverage, -provenance) |>
  knitr::kable(digits = 3, align = "l")
```

Because every task of an occupation carries the same score, the rollup returns
the published occupation score for each occupation with rated core tasks. Chief
Executives have a score in the extract but no tasks in the fixture panel, so
their employment stays in the denominator as unmatched, and coverage is below 1.

## Compare Definitions and Sources

```{r compare-sources}
model_exposure <- onet_import_eloundou(
  task_panel,
  path = eloundou_extract,
  score = "dv_rating_beta",
  measure_id = "eloundou_dv_beta",
  release_version = "30.3"
)

felten_extract <- tempfile(fileext = ".csv")
utils::write.csv(
  data.frame(
    `SOC Code` = c("15-1252", "29-1141", "11-1011"),
    AIOE = c(1.10, -0.30, 0.80),
    check.names = FALSE
  ),
  felten_extract,
  row.names = FALSE
)
aioe <- onet_import_felten_aioe(
  task_panel,
  path = felten_extract,
  release_version = "30.3"
)

aggregate_measure <- function(measure) {
  measure |>
    onet_task_to_occupation(task_ratings = ratings, task_metadata = tasks) |>
    onet_measure_aggregate(weights)
}

list(
  `Human-rated beta` = exposure,
  `Model-rated beta` = model_exposure,
  AIOE = aioe
) |>
  purrr::map(aggregate_measure) |>
  purrr::list_rbind(names_to = "source") |>
  select(source, measure_id, aggregate, employment_coverage_share) |>
  knitr::kable(digits = 3, align = "l")
```

The AIOE adapter joins on the 6-digit SOC code, so every O&#42;NET detail code
under one SOC inherits the same score. The two sources are on different scales:
the beta scores are shares of exposed tasks, while AIOE is a standardized index.
Compare orderings across sources, not levels.

## Occupations without a Score

Tasks whose occupation has no score in the file are dropped with a warning
rather than scored as zero.

```{r unmatched-occupation}
partial_panel <- tibble::tibble(
  onet_soc_code = c("15-1252.00", "51-2092.00"),
  task_id = c("1001", "9001")
)

partial <- onet_import_eloundou(partial_panel, path = eloundou_extract)

partial$data |>
  select(onet_soc_code, task_id, measure_score) |>
  knitr::kable(digits = 3, align = "l")
```

## Reading the Published Files

With no `path`, each adapter downloads the authors' file from a URL pinned to a
specific commit and caches it in the `reference` section of the package cache.
Pinning the digest and an `as_of` label makes a rerun fail loudly if the source
ever changes.

```{r online-import, eval = FALSE}
exposure <- onet_import_eloundou(
  task_panel,
  as_of = "GPTs-are-GPTs commit 0471612",
  expected_sha256 = "<digest recorded on first download>"
)

aioe <- onet_import_felten_aioe(task_panel)

# Remove cached reference downloads.
onet_cache_clear(what = "reference")
```

The Eloundou et al. table is distributed under the MIT License. The AIOE workbook
is provided for research use. Cite the source papers when you use either score.

## References
