---
title: "Why Longitudinal O*NET Analysis Is Hard"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Why Longitudinal O*NET Analysis Is Hard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
helper_paths <- c(
  system.file("vignette-helpers.R", package = "onet2r"),
  "inst/vignette-helpers.R",
  "../inst/vignette-helpers.R",
  "vignette-helpers.R"
)
helper_path <- helper_paths[nzchar(helper_paths) & file.exists(helper_paths)]
if (length(helper_path) == 0) {
  stop("Could not find vignette-helpers.R.", call. = FALSE)
}
source(helper_path[[1]])

archive_base <- onet2r_inst_path("extdata", "onet-mini")
archive_243 <- file.path(archive_base, "db_24_3_text")
archive_251 <- file.path(archive_base, "db_25_1_text")
archive_302 <- file.path(archive_base, "db_30_2_text")
archive_303 <- file.path(archive_base, "db_30_3_text")
```

O&#42;NET is one of the most useful public sources for occupation-level skills,
tasks, abilities, work activities, and work contexts. It is also easy to misuse
as a time series. The most important fact is simple: O&#42;NET was not built as a
longitudinal panel. The Web Services API serves the current release, while
version-over-version analysis has to be reconstructed from the downloadable
release archives [@onet_archive].

That reconstruction is why `onet2r` includes archive, bridge, and reconciliation
helpers. They do not make the historical data automatically causal or
comparable. They make the assumptions visible.

## The 3 Problems Users Must Handle

### 1. The API Is Current-Release First

Most API workflows ask "what does this occupation look like now?" Longitudinal
work asks "what changed between release A and release B?" Those are different
questions. For panels, use archive releases rather than the current-release API.

```{r archive-output}
panel <- onet_panel(
  "Abilities",
  versions = c("30.2", "30.3"),
  scale = "IM",
  archives = c(`30.2` = archive_302, `30.3` = archive_303),
  release_dates = c(`30.2` = "2026-02-01", `30.3` = "2026-05-01")
)

panel |>
  count(release_version, release_date, soc_vintage, domain) |>
  onet_kable()
```

The table above is produced by `onet_panel()`. In a live workflow, the same call
can download archive ZIP files by version. In a package vignette, local archive
paths keep the output reproducible.

### 2. Occupation Codes Change across Taxonomies

O&#42;NET-SOC vintages do not line up one-to-one forever. Occupations split, merge,
appear, and disappear. Hosseinioun et al. [-@hosseinioun2025skill] handle this
by composing adjacent official O&#42;NET-SOC crosswalks, which is the same strategy
used by `onet_crosswalk_bridge()`.

The synthetic fixture below spans the 2010-to-2019 O&#42;NET-SOC seam. One 2010
software occupation maps to 2 2019 occupations, so the bridge is uncertain for
within-occupation comparison.

```{r cross-vintage-panel}
cross_panel <- onet_panel(
  "Abilities",
  versions = c("24.3", "25.1"),
  scale = "IM",
  archives = c(`24.3` = archive_243, `25.1` = archive_251),
  release_dates = c(`24.3` = "2020-08-01", `25.1` = "2020-11-01")
)

bridge_2010_2019 <- tibble::tibble(
  from_vintage = "2010",
  to_vintage = "2019",
  from_onet_soc_code = c("15-1132.00", "15-1132.00", "29-1141.00"),
  to_onet_soc_code = c("15-1252.00", "15-1253.00", "29-1141.00"),
  map_type = c("split", "split", "one_to_one"),
  crosswalk_weight = c(0.5, 0.5, 1)
)

cross_changes <- onet_panel_reconcile(cross_panel, bridge_2010_2019)

cross_changes |>
  select(
    from_onet_soc_code,
    to_onet_soc_code,
    element_name,
    change_type,
    crosswalk_uncertain,
    transition_data,
    safely_comparable
  ) |>
  onet_kable()
```

### 3. A Release-to-Release "Change" May Not Be a True Content Change

Many O&#42;NET archive tables include a source date and domain source. These fields
matter because a value can appear unchanged simply because the occupation was
not re-surveyed, or a value can change without a source-date change because of a
recode or recalculation. Handel [-@handel2016onet] is a useful entry point for
the broader measurement and comparability cautions.

```{r truth-table-from-functions}
changes <- onet_panel_reconcile(
  panel,
  bridge = onet_crosswalk_bridge("2019", "2019")
)

changes |>
  distinct(value_changed, date_changed, change_type, safely_comparable) |>
  arrange(change_type) |>
  onet_kable()
```

```{r comparable-chart, fig.alt = "Horizontal bar chart showing safely comparable and not safely comparable rows in the cross-vintage example."}
comparison_counts <- cross_changes |>
  mutate(comparability = if_else(safely_comparable, "Safe", "Not safe")) |>
  count(comparability, name = "rows")

ggplot2::ggplot(comparison_counts, ggplot2::aes(
  x = comparability,
  y = rows,
  fill = comparability
)) +
  ggplot2::geom_col(width = 0.6, show.legend = FALSE) +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(
    values = c("Safe" = onet2r_colors[["teal"]], "Not safe" = onet2r_colors[["rose"]])
  ) +
  ggplot2::labs(
    title = "How Many Rows Are Safe to Compare?",
    subtitle = "The count comes from the cross-vintage example, not a hand-built table.",
    x = NULL,
    y = "Rows"
  ) +
  onet2r_theme()
```

Those rows are not a hand-built truth table. They are the observed
classifications from the example archive panels. The patterns are the minimum
checks users should make before interpreting any cross-release difference.

### Re-ratings are not independent

Two facts sharpen the carry-forward caution. First, the Skills domain moved
from incumbent to trained-analyst ratings around 2008, a documented
comparability break [@handel2016onet]. Second, for occupations rated a second
time, O&#42;NET's analyst procedures explicitly provide prior ratings to the
analysts [@onet_aoskills_update]. Modern release-over-release changes in
analyst-rated domains are therefore not independent re-measurements; treat
small movements as measurement persistence, not signal. The official
[Longitudinal Data Updates record](https://www.onetcenter.org/dataUpdates.html)
(see `onet_data_updates()`) tells you which occupations were actually re-rated
in each cycle.

## How This Differs from the Standard Task Approach

The canonical task-framework literature often avoids O&#42;NET's
version-over-version variation. Instead, it holds task scores fixed and gets
time variation from changing occupational employment shares [@autor2003skill;
@autordorn2013growth; @autor2013task]. That approach is standard because it is
cleaner for many labor-market questions. `onet2r` enables a complementary route:
directly comparing archived O&#42;NET content after marking taxonomy and
survey-timing problems.

Direct within-occupation change work is closer to what the longitudinal module
does. Consoli et al. [-@consoli2023routinization] combine DOT and O&#42;NET to
study within-occupation task change over a long period. Chen et al.
[-@chen2026ai] report a narrower O&#42;NET version comparison as one stylized fact
inside a paper whose main data source is online vacancies.

## Official Resources to Check

- The O&#42;NET Database Releases Archive is the source for release ZIP files
  [@onet_archive].
- The O&#42;NET Data Dictionary documents table structure, date fields, source
  fields, and historical content changes [@onet_dictionary].
- The O&#42;NET-SOC taxonomy files provide the official adjacent crosswalks used by
  the bridge layer [@onet_crosswalks].
- BLS provides a methodological overview for mapping Employment Projections and
  O&#42;NET data, including split, merge, and imputation issues that motivate
  future employment-weighted bridge options [@bls2021mapping].

## Adjacent Tools to Verify before Citing

Several tools map O&#42;NET to other classification systems rather than across
O&#42;NET's own versions. They may be useful, but they are intentionally not in
`inst/REFERENCES.bib` until verified to citation standard:

- Institute for Structural Research occupational task crosswalk files.
- `occupationcross`, reported as an R package for occupational reclassification.
- The Dais NOC-to-O&#42;NET crosswalk for Canada.
- European Commission ESCO-O&#42;NET crosswalks.

## References
