---
title: "Why Longitudinal O*NET Analysis Is Hard"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Why Longitudinal O*NET Analysis Is Hard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(onet2r)
library(tibble)
```

O*NET is one of the most useful public sources for occupation-level skills,
tasks, abilities, work activities, and work contexts. It is also easy to misuse
as a time series. The most important fact is simple: O*NET was not built as a
longitudinal panel. The Web Services API serves the current release, while
version-over-version analysis has to be reconstructed from the downloadable
release archives [@onet_archive].

That reconstruction is the reason `onet2r` includes archive, bridge, and
reconciliation helpers. They do not make the historical data automatically
causal or comparable. They make the assumptions visible.

## The 3 Problems Users Must Handle

### 1. The API Is Current-Release First

Most API workflows ask "what does this occupation look like now?" Longitudinal
work asks "what changed between release A and release B?" Those are different
questions. For panels, use the archive functions.

```{r archive-schema}
release_example <- tibble(
  version = c("30.3", "30.2", "30.1"),
  release_date = as.Date(c("2026-05-01", "2026-02-01", "2025-11-01")),
  soc_vintage = c("2019", "2019", "2019"),
  text_url = c(
    "https://www.onetcenter.org/dl_files/database/db_30_3_text.zip",
    "https://www.onetcenter.org/dl_files/database/db_30_2_text.zip",
    "https://www.onetcenter.org/dl_files/database/db_30_1_text.zip"
  )
)

head(release_example, 3)
```

### 2. Occupation Codes Change across Taxonomies

O*NET-SOC vintages do not line up one-to-one forever. Occupations split, merge,
appear, and disappear. Hosseinioun et al. [-@hosseinioun2025skill] handle this
by composing adjacent official O*NET-SOC crosswalks, which is the same strategy
used by `onet_crosswalk_bridge()`.

```{r crosswalk-example}
bridge_example <- tibble(
  from_vintage = "2010",
  to_vintage = "2019",
  from_soc_code = c("15-1132", "15-1132", "29-1141"),
  to_soc_code = c("15-1252", "15-1253", "29-1141"),
  map_type = c("split", "split", "one_to_one"),
  crosswalk_weight = c(0.5, 0.5, 1)
)

head(bridge_example, 3)
```

Rows marked `split` or `merge` are not thrown away. They are flagged as
crosswalk-uncertain during reconciliation so users can decide whether to keep,
weight, or exclude them.

### 3. A Release-to-Release "Change" May Not Be a True Content Change

Many O*NET archive tables include a source date and domain source. These fields
matter because a value can appear unchanged simply because the occupation was
not re-surveyed, or a value can change without a source-date change because of a
recode or recalculation. Handel [-@handel2016onet] is a useful entry point for
the broader measurement and comparability cautions.

```{r truth-table}
tibble(
  value_changed = c(FALSE, TRUE, FALSE, TRUE),
  source_date_changed = c(FALSE, TRUE, TRUE, FALSE),
  change_type = c(
    "stale_carryforward",
    "real_update",
    "resampled_stable",
    "recode_or_recalc_flag"
  ),
  comparable_before_overrides = c(TRUE, TRUE, TRUE, FALSE)
)
```

## How This Differs from the Standard Task Approach

The canonical task-framework literature often avoids O*NET's
version-over-version variation. Instead, it holds task scores fixed and gets
time variation from changing occupational employment shares [@autor2003skill;
@autordorn2013growth; @autor2013task]. That approach is standard because it is
cleaner for many labor-market questions. `onet2r` enables a complementary route:
directly comparing archived O*NET content after marking taxonomy and
survey-timing problems.

Direct within-occupation change work is closer to what the longitudinal module
does. Consoli et al. [-@consoli2023routinization] combine DOT and O*NET to study
within-occupation task change over a long period. Chen et al. [-@chen2026ai]
report a narrower O*NET version comparison as one stylized fact inside a paper
whose main data source is online vacancies.

## Official Resources to Check

- The O*NET Database Releases Archive is the source for release ZIP files
  [@onet_archive].
- The O*NET Data Dictionary documents table structure, date fields, source
  fields, and historical content changes [@onet_dictionary].
- The O*NET-SOC taxonomy files provide the official adjacent crosswalks used by
  the bridge layer [@onet_crosswalks].
- BLS provides a methodological overview for mapping Employment Projections and
  O*NET data, including split, merge, and imputation issues that motivate future
  employment-weighted bridge options [@bls2021mapping].

## Adjacent Tools to Verify before Citing

Several tools map O*NET to other classification systems rather than across
O*NET's own versions. They may be useful, but they are intentionally not in
`inst/REFERENCES.bib` until verified to citation standard:

- Institute for Structural Research occupational task crosswalk files.
- `occupationcross`, reported as an R package for occupational reclassification.
- The Dais NOC-to-O*NET crosswalk for Canada.
- European Commission ESCO-O*NET crosswalks.

## References
