---
title: "Longitudinal O*NET Archive Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal O*NET Archive Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
helper_paths <- c(
  "inst/vignette-helpers.R",
  "vignette-helpers.R",
  system.file("vignette-helpers.R", package = "onet2r")
)
helper_path <- helper_paths[nzchar(helper_paths) & file.exists(helper_paths)][[1]]
source(helper_path)

example_archive <- function(version) {
  system.file(
    "extdata",
    "onet-mini",
    paste0("db_", gsub(".", "_", version, fixed = TRUE), "_text"),
    package = "onet2r"
  )
}
```

Read `vignette("longitudinal-onet-background", package = "onet2r")` before
interpreting changes. O&#42;NET was not designed as a longitudinal panel, so the
practical job is not only to compute differences. It is to separate real
descriptor updates from carryforwards, re-surveys with stable values, recodes,
transition rows, and taxonomy seams.

This walkthrough uses packaged archive-format text files so every displayed
table comes from `onet2r` functions while the vignette remains CRAN-safe and
does not require network access.

## A Practical Question

Suppose you want to know whether selected cognitive ability ratings changed
between O&#42;NET releases for occupations that matter in workforce planning:
software developers, registered nurses, chief executives, and retail
supervisors. A naive `new - old` difference is not enough, because unchanged
values may be carryforwards and changed values may reflect a recode rather than
a fresh survey update.

## Read One Archive Table

`onet_archive_read()` reads an O&#42;NET text archive table and normalizes it into
one panel schema.

```{r read-one-archive}
abilities_303 <- onet_archive_read(
  "30.3",
  "Abilities",
  path = example_archive("30.3"),
  release_date = "2026-05-01"
)

abilities_303 |>
  select(
    release_version,
    onet_soc_code,
    soc_code,
    element_id,
    element_name,
    data_value,
    source_date,
    domain_source
  ) |>
  head(8) |>
  onet_kable()
```

The output has one row per occupation, element, scale, and release. The
important longitudinal fields are `release_version`, `soc_vintage`,
`onet_soc_code`, `data_value`, `source_date`, and `domain_source`.

## Assemble a Same-Vintage Panel

`onet_panel()` applies the same reader across releases and row-binds the
normalized outputs.

```{r build-panel}
same_vintage_archives <- c(
  `30.2` = example_archive("30.2"),
  `30.3` = example_archive("30.3")
)
same_vintage_dates <- c(`30.2` = "2026-02-01", `30.3` = "2026-05-01")

panel <- onet_panel(
  "Abilities",
  versions = c("30.2", "30.3"),
  scale = "IM",
  archives = same_vintage_archives,
  release_dates = same_vintage_dates
)

panel |>
  arrange(onet_soc_code, element_id, release_version) |>
  select(
    release_version,
    onet_soc_code,
    element_name,
    data_value,
    source_date,
    domain_source
  ) |>
  head(10) |>
  onet_kable()
```

This is the table you would save as your audit trail before doing any modeling.
It records the exact versions, values, and source dates used in the comparison.

## Reconcile Adjacent Releases

`onet_panel_reconcile()` compares adjacent releases and classifies each matched
occupation-element-scale pair.

```{r reconcile}
changes <- onet_panel_reconcile(panel, onet_crosswalk_bridge("2019", "2019"))

changes |>
  select(
    from_soc_code,
    to_soc_code,
    element_name,
    from_value,
    to_value,
    value_change,
    from_source_date,
    to_source_date,
    change_type,
    method_break,
    safely_comparable
  ) |>
  arrange(desc(abs(value_change))) |>
  head(10) |>
  onet_kable()
```

Read `change_type` before interpreting `value_change`.

| Value changed? | Source date changed? | Classification | Interpretation |
| --- | --- | --- | --- |
| no | no | `stale_carryforward` | The release likely carried forward the prior value. |
| yes | yes | `real_update` | The value changed with a new source date. |
| no | yes | `resampled_stable` | The occupation appears updated, but the score stayed stable. |
| yes | no | `recode_or_recalc_flag` | Treat cautiously because the value changed without a new source date. |

```{r changed-rows}
changes |>
  filter(value_changed) |>
  mutate(abs_change = abs(value_change)) |>
  arrange(safely_comparable, desc(abs_change)) |>
  select(
    to_soc_code,
    element_name,
    from_value,
    to_value,
    value_change,
    change_type,
    method_break,
    safely_comparable
  ) |>
  onet_kable()
```

## Cross a Taxonomy Seam

Same-vintage comparisons are the easy case. The next example uses the bundled
2010-vintage and 2019-vintage fixtures. The bridge is intentionally tiny, but it
shows the real problem: a 2010 occupation can split into more than one 2019
occupation.

```{r cross-vintage}
cross_archives <- c(
  `24.3` = example_archive("24.3"),
  `25.1` = example_archive("25.1")
)
cross_dates <- c(`24.3` = "2020-08-01", `25.1` = "2020-11-01")

cross_panel <- onet_panel(
  "Abilities",
  versions = c("24.3", "25.1"),
  scale = "IM",
  archives = cross_archives,
  release_dates = cross_dates
)

bridge_2010_2019 <- tibble::tibble(
  from_vintage = "2010",
  to_vintage = "2019",
  from_onet_soc_code = c("15-1132.00", "15-1132.00", "29-1141.00"),
  to_onet_soc_code = c("15-1252.00", "15-1253.00", "29-1141.00"),
  map_type = c("split", "split", "one_to_one"),
  crosswalk_weight = c(0.5, 0.5, 1)
)

cross_changes <- onet_panel_reconcile(cross_panel, bridge_2010_2019)

cross_changes |>
  select(
    from_onet_soc_code,
    to_onet_soc_code,
    element_name,
    change_type,
    crosswalk_uncertain,
    transition_data,
    safely_comparable
  ) |>
  onet_kable()
```

```{r summary-chart, fig.alt = "Dumbbell chart of cross-vintage ability changes, with safe comparisons separated from unsafe comparisons."}
plot_changes <- cross_changes |>
  filter(!is.na(from_value), !is.na(to_value)) |>
  mutate(
    comparison = paste(to_onet_soc_code, element_name, sep = ": "),
    comparability = if_else(safely_comparable, "Safe", "Not safe")
  )

ggplot2::ggplot(plot_changes, ggplot2::aes(y = comparison)) +
  ggplot2::geom_segment(
    ggplot2::aes(
      x = from_value,
      xend = to_value,
      yend = comparison,
      color = comparability
    ),
    linewidth = 1.1
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = from_value),
    color = onet2r_colors[["slate"]],
    size = 2.8
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = to_value, color = comparability),
    size = 3.2
  ) +
  onet2r_discrete_color(name = "Comparability") +
  ggplot2::labs(
    title = "A Cross-Vintage Difference Is Not Automatically Comparable",
    subtitle = "Start and end points come from the bundled archive fixtures.",
    x = "Importance rating",
    y = NULL
  ) +
  onet2r_theme()
```

## Summarize the Panel

`onet_change_summary()` gives a compact audit of the reconciliation result. The
job-family view is useful for spotting whether one part of the SOC taxonomy is
driving the apparent change.

```{r summary}
onet_change_summary(changes, by = "job_family") |>
  onet_kable()
```

## Recommended Workflow for Real Archive Work

1. Use `onet_releases()` to identify release versions and O&#42;NET-SOC vintages.
2. Build a panel with `onet_panel()` for one domain at a time, such as
   `"Abilities"`, `"Skills"`, or `"Work Activities"`.
3. Build a bridge with `onet_crosswalk_bridge()` when releases use different
   O&#42;NET-SOC vintages.
4. Reconcile with `onet_panel_reconcile()`.
5. Filter, weight, or model only after checking `change_type`, `method_break`,
   `crosswalk_uncertain`, and `safely_comparable`.
