---
title: "Longitudinal O*NET Archive Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal O*NET Archive Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(onet2r)
library(dplyr)
```

Read `vignette("longitudinal-onet-background", package = "onet2r")` before
interpreting changes. O&#42;NET was not designed as a longitudinal panel, so the
practical job is not only to compute differences. It is to separate real
descriptor updates from carryforwards, re-surveys with stable values, recodes,
and taxonomy seams.

This walkthrough uses packaged archive-format text files so every displayed
table comes from `onet2r` functions while the vignette remains CRAN-safe and
does not require network access. In a real project, replace the local paths with
downloaded O&#42;NET archive versions.

```{r local-archives}
example_archive <- function(version) {
  system.file(
    "extdata",
    "onet-mini",
    paste0("db_", gsub(".", "_", version, fixed = TRUE), "_text"),
    package = "onet2r"
  )
}

archives <- c(
  `30.2` = example_archive("30.2"),
  `30.3` = example_archive("30.3")
)

release_dates <- c(`30.2` = "2026-02-01", `30.3` = "2026-05-01")

archives
```

## A Practical Question

Suppose you want to know whether selected cognitive ability ratings changed
between two O&#42;NET releases for occupations that matter in workforce planning:
software developers, registered nurses, chief executives, and retail
supervisors. A naive `new - old` difference is not enough, because unchanged
values may be carryforwards and changed values may reflect a recode rather than
a fresh survey update.

## Read One Archive Table

`onet_archive_read()` reads an O&#42;NET text archive table and normalizes it into
one panel schema. The local `path` argument is useful when you have downloaded
archives manually or when you want a reproducible, offline workflow.

```{r read-one-archive}
abilities_303 <- onet_archive_read(
  "30.3",
  "Abilities",
  path = archives[["30.3"]],
  release_date = release_dates[["30.3"]]
)

head(abilities_303, 8)
```

The output has one row per occupation, element, scale, and release. The
important longitudinal fields are:

- `release_version` and `release_date`, which identify the archive release;
- `soc_vintage`, `onet_soc_code`, and `soc_code`, which identify the taxonomy;
- `data_value`, the descriptor score;
- `source_date` and `domain_source`, which help distinguish fresh updates from
  carryforwards or method changes.

## Assemble a Two-Release Panel

`onet_panel()` applies the same reader across releases and row-binds the
normalized outputs.

```{r build-panel}
panel <- onet_panel(
  "Abilities",
  versions = c("30.2", "30.3"),
  scale = "IM",
  archives = archives,
  release_dates = release_dates
)

panel |>
  arrange(onet_soc_code, element_id, release_version) |>
  head(10)
```

This is the table you would save as your audit trail before doing any modeling.
It records the exact versions, values, and source dates used in the comparison.

## Bridge Taxonomy Vintages

The example releases both use the 2019 O&#42;NET-SOC vintage, so no crosswalk is
needed. Calling `onet_crosswalk_bridge("2019", "2019")` returns the empty bridge
schema that `onet_panel_reconcile()` can use for same-vintage comparisons. For
real cross-vintage work, use `onet_crosswalk_bridge()` to compose official
adjacent O&#42;NET-SOC crosswalks.

```{r bridge}
bridge <- onet_crosswalk_bridge("2019", "2019")
bridge
```

## Reconcile Adjacent Releases

`onet_panel_reconcile()` compares adjacent releases and classifies each matched
occupation-element-scale pair.

```{r reconcile}
changes <- onet_panel_reconcile(panel, bridge)

changes |>
  select(
    from_soc_code,
    to_soc_code,
    element_name,
    from_value,
    to_value,
    value_change,
    from_source_date,
    to_source_date,
    change_type,
    method_break,
    safely_comparable
  ) |>
  arrange(desc(abs(value_change))) |>
  head(10)
```

Read the `change_type` column before interpreting `value_change`:

| Value changed? | Source date changed? | Classification | Interpretation |
| --- | --- | --- | --- |
| no | no | `stale_carryforward` | The release likely carried forward the prior value. |
| yes | yes | `real_update` | The value changed with a new source date. |
| no | yes | `resampled_stable` | The occupation appears updated, but the score stayed stable. |
| yes | no | `recode_or_recalc_flag` | Treat cautiously; the value changed without a new source date. |

`method_break` and `crosswalk_uncertain` are additional overrides. A row is
`safely_comparable` only when the truth-table result is comparable and neither
override is present.

## Answer the Practical Question

For this small example, the most useful review table is not every pair. It is
the subset where values changed, sorted by whether the change is safe to compare.

```{r changed-rows}
changes |>
  filter(value_changed) |>
  mutate(abs_change = abs(value_change)) |>
  arrange(safely_comparable, desc(abs_change)) |>
  select(
    to_soc_code,
    element_name,
    from_value,
    to_value,
    value_change,
    change_type,
    method_break,
    safely_comparable
  )
```

This is the table a researcher should inspect before making a claim such as
"software developers require more oral comprehension than in the prior release."
The comparison is strongest when the value changed, the source date changed, and
no method or taxonomy warning is present.

## Summarize the Panel

`onet_change_summary()` gives a compact audit of the reconciliation result. The
job-family view is useful for spotting whether one part of the SOC taxonomy is
driving the apparent change.

```{r summary}
onet_change_summary(changes, by = "job_family")
```

## Recommended Workflow for Real Archive Work

1. Use `onet_releases()` to identify release versions and O&#42;NET-SOC vintages.
2. Build a panel with `onet_panel()` for one domain at a time, such as
   `"Abilities"`, `"Skills"`, or `"Work Activities"`.
3. Build a bridge with `onet_crosswalk_bridge()` when releases use different
   O&#42;NET-SOC vintages.
4. Reconcile with `onet_panel_reconcile()`.
5. Filter, weight, or model only after checking `change_type`, `method_break`,
   `crosswalk_uncertain`, and `safely_comparable`.
