---
title: "Longitudinal O*NET archive analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal O*NET archive analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(onet2r)
library(tibble)
```

The longitudinal helpers support three related workflows:

1. Discover available O*NET database releases.
2. Read descriptor tables from archived text files into a consistent panel.
3. Reconcile adjacent releases and separate real updates from carryforwards,
   resampling, recodes, and crosswalk uncertainty.

Live archive calls are guarded because they download files from O*NET. In an
interactive session, start by listing releases and reading a descriptor table:

```{r live-archive, eval = FALSE}
releases <- onet_releases()
head(releases)

abilities <- onet_archive_read("30.3", "Abilities")
head(abilities)

panel <- onet_panel("Abilities", versions = c("30.2", "30.3"), scale = "IM")
```

For reproducible examples and tests, the reconciliation step can use any tibble
with the same schema returned by `onet_panel()`. This tiny panel contains four
occupation-element pairs across two releases. Each pair demonstrates one branch
of the change classification truth table.

```{r tiny-panel}
panel <- tibble(
  release_version = rep(c("1.0", "2.0"), each = 4),
  release_date = rep(as.Date(c("2020-01-01", "2021-01-01")), each = 4),
  soc_vintage = "2019",
  domain = "Abilities",
  onet_soc_code = rep(
    c("11-1011.00", "15-1252.00", "29-1141.00", "41-1011.00"),
    2
  ),
  soc_code = rep(c("11-1011", "15-1252", "29-1141", "41-1011"), 2),
  title = "Occupation",
  element_id = "1.A.1.a.1",
  element_name = "Oral Comprehension",
  scale_id = "IM",
  data_value = c(4, 4, 4, 4, 4, 5, 4, 5),
  source_date = c(
    as.Date(c("2020-01-01", "2020-01-01", "2020-01-01", "2020-01-01")),
    as.Date(c("2020-01-01", "2021-01-01", "2021-01-01", "2020-01-01"))
  ),
  domain_source = "Analyst"
)

bridge <- tibble(
  from_vintage = "2019",
  to_vintage = "2019",
  from_soc_code = c("11-1011", "15-1252", "29-1141", "41-1011"),
  to_soc_code = c("11-1011", "15-1252", "29-1141", "41-1011"),
  from_title = "Occupation",
  to_title = "Occupation",
  step_count = 0L,
  map_type = "one_to_one",
  crosswalk_weight = 1
)

changes <- onet_panel_reconcile(panel, bridge)
changes[, c(
  "from_soc_code",
  "to_soc_code",
  "value_change",
  "date_changed",
  "change_type",
  "safely_comparable"
)]
```

The truth table is:

| value changed | source date changed | `change_type` | initially comparable |
| --- | --- | --- | --- |
| no | no | `stale_carryforward` | yes |
| yes | yes | `real_update` | yes |
| no | yes | `resampled_stable` | yes |
| yes | no | `recode_or_recalc_flag` | no |

`method_break` and `crosswalk_uncertain` are override flags. A row is safely
comparable only when the truth-table result is comparable and neither override
flag is present.

```{r summary}
onet_change_summary(changes, by = "job_family")
```

When comparing across taxonomy vintages, create a bridge first:

```{r bridge, eval = FALSE}
bridge_2010_2019 <- onet_crosswalk_bridge("2010", "2019")
table(bridge_2010_2019$map_type)
```

Rows with `map_type` equal to `split` or `merge` are retained, but
`crosswalk_uncertain` is set to `TRUE` during reconciliation. This lets users
keep the data while avoiding overconfident longitudinal claims.
