---
title: "Longitudinal O*NET Archive Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal O*NET Archive Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(onet2r)
library(tibble)
```

Read `vignette("longitudinal-onet-background", package = "onet2r")` before
interpreting changes. This article focuses on the mechanics: assemble a panel,
bridge taxonomy vintages, reconcile adjacent releases, and summarize the result.

## Start with the Release Table

Live release discovery reads the public O*NET Resource Center archive page. The
package function returns one row per downloadable text archive. The example
below uses a small representative table so the article builds without network
access.

```{r releases}
releases <- tibble(
  version = c("30.3", "30.2", "30.1"),
  release_date = as.Date(c("2026-05-01", "2026-02-01", "2025-11-01")),
  year = c(2026L, 2026L, 2025L),
  month = c("May", "February", "November"),
  soc_vintage = c("2019", "2019", "2019"),
  text_url = c(
    "https://www.onetcenter.org/dl_files/database/db_30_3_text.zip",
    "https://www.onetcenter.org/dl_files/database/db_30_2_text.zip",
    "https://www.onetcenter.org/dl_files/database/db_30_1_text.zip"
  ),
  dictionary_url = c(
    "https://www.onetcenter.org/dictionary/30.3/excel/",
    "https://www.onetcenter.org/dictionary/30.2/excel/",
    "https://www.onetcenter.org/dictionary/30.1/excel/"
  )
)

head(releases, 3)
```

In an interactive workflow, call `onet_releases()` and then choose the versions
you want to compare.

## Assemble a Descriptor Panel

`onet_archive_read()` normalizes descriptor-like archive tables into a long
schema. `onet_panel()` row-binds the same table across releases. This example
uses a tiny panel with the same columns.

```{r panel}
panel <- tibble(
  release_version = rep(c("1.0", "2.0"), each = 4),
  release_date = rep(as.Date(c("2020-01-01", "2021-01-01")), each = 4),
  soc_vintage = "2019",
  domain = "Abilities",
  onet_soc_code = rep(
    c("11-1011.00", "15-1252.00", "29-1141.00", "41-1011.00"),
    2
  ),
  soc_code = rep(c("11-1011", "15-1252", "29-1141", "41-1011"), 2),
  title = "Occupation",
  element_id = "1.A.1.a.1",
  element_name = "Oral Comprehension",
  scale_id = "IM",
  data_value = c(4, 4, 4, 4, 4, 5, 4, 5),
  source_date = c(
    as.Date(c("2020-01-01", "2020-01-01", "2020-01-01", "2020-01-01")),
    as.Date(c("2020-01-01", "2021-01-01", "2021-01-01", "2020-01-01"))
  ),
  domain_source = "Analyst"
)

head(panel, 6)
```

For live data, use the same pattern with archive helpers:
`onet_panel("Abilities", versions = c("30.2", "30.3"), scale = "IM")`.

## Bridge Taxonomy Vintages

When both releases use the same SOC vintage, the bridge is one-to-one. Across
vintages, `onet_crosswalk_bridge()` composes official adjacent crosswalks and
labels split or merge rows.

```{r bridge}
bridge <- tibble(
  from_vintage = "2019",
  to_vintage = "2019",
  from_soc_code = c("11-1011", "15-1252", "29-1141", "41-1011"),
  to_soc_code = c("11-1011", "15-1252", "29-1141", "41-1011"),
  from_title = "Occupation",
  to_title = "Occupation",
  step_count = 0L,
  map_type = "one_to_one",
  crosswalk_weight = 1
)

head(bridge, 6)
```

## Reconcile Adjacent Releases

`onet_panel_reconcile()` compares adjacent releases and classifies each matched
occupation-element-scale pair.

```{r reconcile}
changes <- onet_panel_reconcile(panel, bridge)

head(changes[, c(
  "from_soc_code",
  "to_soc_code",
  "from_value",
  "to_value",
  "value_change",
  "date_changed",
  "change_type",
  "safely_comparable"
)], 8)
```

The change classification is:

| Value Changed | Source Date Changed | Change Type | Initially Comparable |
| --- | --- | --- | --- |
| no | no | `stale_carryforward` | yes |
| yes | yes | `real_update` | yes |
| no | yes | `resampled_stable` | yes |
| yes | no | `recode_or_recalc_flag` | no |

`method_break` and `crosswalk_uncertain` are overrides. A row is safely
comparable only when the truth-table result is comparable and neither override
is present.

## Summarize Changes

Summaries give a quick audit of how much of the panel is safely comparable and
how often each change pattern appears.

```{r summary}
summary <- onet_change_summary(changes, by = "job_family")
head(summary, 10)
```

## Recommended Workflow

1. Pick a domain table such as `Abilities`, `Skills`, or `Work Activities`.
2. Use `onet_releases()` to identify archive versions and SOC vintages.
3. Build a panel with `onet_panel()`.
4. Build a bridge with `onet_crosswalk_bridge()`.
5. Reconcile with `onet_panel_reconcile()`.
6. Inspect `method_break`, `crosswalk_uncertain`, and `safely_comparable` before
   drawing substantive conclusions.
