---
title: "Reproducible user measures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducible user measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(tibble.print_min = 4, tibble.print_max = 8)
options(
  cli.num_colors = 1,
  crayon.enabled = FALSE,
  pillar.bold = FALSE,
  pillar.subtle = FALSE
)
```

```{r setup}
library(onet2r)
library(dplyr)
```

O&#42;NET users often arrive with their own score: an AI exposure measure, a task
classification, a skill index, or a hand-coded construct. `onet2r` does not try
to decide which score is correct. It helps with the parts around the score:
checking keys, using versioned O&#42;NET files, aggregating tasks to occupations,
adding employment weights, and recording enough provenance for someone else to
reproduce the number.

This article uses the small archive fixtures shipped with the package. The
numbers are stylized, but every table below is produced by package functions.

## Read task files from a pinned release

```{r read-task-fixtures}
archive_dir <- system.file(
  "extdata",
  "onet-mini",
  "db_30_3_text",
  package = "onet2r"
)
source_archive_dir <- file.path("inst", "extdata", "onet-mini", "db_30_3_text")
if (
  !file.exists(file.path(archive_dir, "task_statements.txt")) &&
    dir.exists(source_archive_dir)
) {
  archive_dir <- source_archive_dir
}

tasks <- onet_archive_read(
  "30.3",
  "Task Statements",
  path = archive_dir,
  release_date = "2026-05-01"
)
task_ratings <- onet_archive_read(
  "30.3",
  "Task Ratings",
  path = archive_dir,
  release_date = "2026-05-01"
)

tasks |>
  select(onet_soc_code, task_id, task_type, task) |>
  head() |>
  print(width = Inf)
```

`Task Statements` carries task ids and Core or Supplemental labels. `Task
Ratings` carries the relevance, importance, and frequency ratings used to
aggregate task-level scores.

```{r task-ratings-preview}
task_ratings |>
  select(onet_soc_code, task_id, scale_id, data_value, recommend_suppress) |>
  head() |>
  print(width = Inf)
```

## Validate a user-supplied measure

Here the user brings a three-task score. It could come from a model, a survey,
or manual coding. The constructor checks that the task ids match the selected
O&#42;NET release universe.

```{r construct-measure}
task_scores <- tibble::tibble(
  task_id = c("1001", "1002", "2001"),
  score = c(0.80, 0.40, 0.20)
)

measure <- onet_measure(
  task_scores,
  key = "task_id",
  score = "score",
  key_type = "task",
  universe = tasks$task_id,
  measure_id = "stylized_task_score",
  measure_name = "Stylized task score",
  release_version = "30.3"
)

onet_measure_coverage(measure)
```

The object also stores unmatched keys explicitly.

```{r unmatched}
if (nrow(measure$unmatched) == 0) {
  tibble::tibble(status = "No unmatched task ids")
} else {
  measure$unmatched
}
```

## Roll tasks to occupations

The next step is mechanical. We use O&#42;NET task relevance ratings as weights
and restrict the calculation to Core tasks.

```{r task-rollup}
occupation_scores <- onet_task_to_occupation(
  measure,
  task_ratings = task_ratings,
  task_metadata = tasks,
  weight_scale = "RT",
  include_supplemental = FALSE
)

occupation_scores
```

Changing the Core-only rule is a plumbing choice, not a change to the user's
score. It belongs in provenance and robustness checks.

```{r task-rollup-all}
occupation_scores_all <- onet_task_to_occupation(
  measure,
  task_ratings = task_ratings,
  task_metadata = tasks,
  weight_scale = "RT",
  include_supplemental = TRUE
)

occupation_scores_all
```

## Add employment weights

OEWS files are SOC-level. The weight helper makes the source vintage explicit
and returns a single shape used downstream.

```{r weights}
oews_sample <- onet_oews_national(
  path = system.file("extdata", "oews-national-sample.csv", package = "onet2r")
)

weights <- onet_weight_panel_oews(oews_sample, year = 2024)

weights |>
  print(width = Inf)
```

## Aggregate and inspect provenance

```{r aggregate}
national <- onet_measure_aggregate(
  occupation_scores,
  weights,
  measure_id = "stylized_task_score"
)

national |>
  print(width = Inf)
attr(national, "provenance")
```

## Check sensitivity to plumbing choices

```{r robustness}
national_all_tasks <- onet_measure_aggregate(
  occupation_scores_all,
  weights,
  measure_id = "stylized_task_score"
)

diagnostic <- onet_robustness_diagnostic(
  tibble::tibble(
    scenario = c("core_tasks", "core_plus_supplemental"),
    aggregate = c(national$aggregate, national_all_tasks$aggregate)
  )
)

diagnostic |>
  print(width = Inf)
```

The diagnostic does not say which task score is right. It shows how much the
headline number moves when the non-substantive plumbing changes.
