---
title: "Reproducible User Measures"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Reproducible User Measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
options(
  cli.num_colors = 1,
  crayon.enabled = FALSE,
  pillar.bold = FALSE,
  pillar.subtle = FALSE
)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
helper_paths <- c(
  system.file("vignette-helpers.R", package = "onet2r"),
  "inst/vignette-helpers.R",
  "../inst/vignette-helpers.R",
  "vignette-helpers.R"
)
helper_path <- helper_paths[nzchar(helper_paths) & file.exists(helper_paths)]
if (length(helper_path) == 0) {
  stop("Could not find vignette-helpers.R.", call. = FALSE)
}
source(helper_path[[1]])

archive_dir <- onet2r_inst_path("extdata", "onet-mini", "db_30_3_text")
```

O&#42;NET users often arrive with their own score: an exposure measure, a task
classification, a skill index, or a hand-coded construct. `onet2r` does not try
to decide which score is correct. It helps with the parts around the score:
checking keys, using versioned O&#42;NET files, aggregating tasks to occupations,
adding employment weights, and recording enough provenance for someone else to
reproduce the number.

This article uses the small archive fixtures shipped with the package. The task
score is stylized and should not be interpreted as a real exposure measure.
Every displayed table is produced by package functions.

## Read Task Files from a Pinned Release

```{r read-task-fixtures}
tasks <- onet_archive_read(
  "30.3",
  "Task Statements",
  path = archive_dir,
  release_date = "2026-05-01"
)
task_ratings <- onet_archive_read(
  "30.3",
  "Task Ratings",
  path = archive_dir,
  release_date = "2026-05-01"
)

tasks |>
  select(onet_soc_code, task_id, task_type, task) |>
  head() |>
  onet_kable()
```

`Task Statements` carries task ids and Core or Supplemental labels. `Task
Ratings` carries relevance and importance ratings used to aggregate task-level
scores.

A practical note on choosing between the Importance (1-5) and Level (0-7)
scales: they are empirically near-redundant. Handel [@handel2016onet] reports
a mean within-descriptor correlation of 0.92 across 130,249 ratings, with
roughly 19% of correlations at 0.98 or above. A measure built from IM and one
built from LV will usually rank occupations almost identically, so treat an
IM-vs-LV sensitivity check as cheap insurance, not as two independent measures.

```{r task-ratings-preview}
task_ratings |>
  select(onet_soc_code, task_id, scale_id, scale_name, data_value, recommend_suppress) |>
  head() |>
  onet_kable()
```

## Validate a User-Supplied Measure

Here the user brings a 3-task score. It could come from a model, a survey, or
manual coding. The constructor checks that the task ids match the selected
O&#42;NET release universe.

```{r construct-measure}
task_scores <- tibble::tibble(
  task_id = c("1001", "1002", "2001"),
  score = c(0.80, 0.40, 0.20)
)

measure <- onet_measure(
  task_scores,
  key = "task_id",
  score = "score",
  key_type = "task",
  universe = tasks$task_id,
  measure_id = "stylized_task_score",
  measure_name = "Stylized task score",
  release_version = "30.3"
)

onet_coverage(measure) |>
  onet_kable()
```

The object also stores unmatched keys explicitly.

```{r unmatched}
unmatched <- if (nrow(measure$unmatched) == 0) {
  tibble::tibble(status = "No unmatched task ids")
} else {
  measure$unmatched
}

unmatched |>
  onet_kable()
```

## Roll Tasks to Occupations

The next step is mechanical. We use O&#42;NET task relevance ratings as weights
and restrict the calculation to Core tasks.

```{r task-rollup}
occupation_scores <- onet_task_to_occupation(
  measure,
  task_ratings = task_ratings,
  task_metadata = tasks,
  weight_scale = "RT",
  include_supplemental = FALSE
)

occupation_scores |>
  select(onet_soc_code, soc_code, n_tasks, total_task_weight, measure_score) |>
  onet_kable()
```

Changing the Core-only rule is a plumbing choice, not a change to the user's
score. It belongs in provenance and sensitivity checks.

```{r task-rollup-all}
occupation_scores_all <- onet_task_to_occupation(
  measure,
  task_ratings = task_ratings,
  task_metadata = tasks,
  weight_scale = "RT",
  include_supplemental = TRUE
)

occupation_scores_all |>
  select(onet_soc_code, soc_code, n_tasks, total_task_weight, measure_score) |>
  onet_kable()
```

## Add Employment Weights

OEWS files are SOC-level. The weight helper makes the source vintage explicit
and returns a single shape used downstream.

```{r weights}
oews_sample <- onet_oews_national(
  path = onet2r_inst_path("extdata", "oews-national-sample.csv")
)

weights <- onet_weight_panel_oews(oews_sample, year = 2024)

weights |>
  onet_kable()
```

## Aggregate and Inspect Provenance

```{r aggregate}
national <- onet_measure_aggregate(
  occupation_scores,
  weights,
  measure_id = "stylized_task_score"
)

national |>
  select(-coverage, -provenance) |>
  onet_kable()

onet_provenance(national) |>
  onet_kable()
```

## Check Sensitivity to Plumbing Choices

```{r robustness}
diagnostic <- onet_measure_sensitivity(
  measure,
  weight_panels = weights,
  task_ratings = task_ratings,
  task_metadata = tasks,
  include_supplemental = c(FALSE, TRUE)
)

diagnostic |>
  select(scenario, aggregate, employment_coverage_share, movement, movement_percent) |>
  onet_kable()
```

```{r robustness-chart, fig.alt = "Horizontal dot chart comparing core-only and core-plus-supplemental stylized task aggregates."}
plot_diagnostic <- diagnostic |>
  mutate(
    rule = if_else(
      include_supplemental,
      "Core + supplemental tasks",
      "Core tasks only"
    )
  )

ggplot2::ggplot(plot_diagnostic, ggplot2::aes(
  x = aggregate,
  y = stats::reorder(rule, aggregate)
)) +
  ggplot2::geom_vline(
    xintercept = plot_diagnostic$baseline_aggregate[[1]],
    color = onet2r_colors[["slate"]],
    linetype = "dashed"
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(x = 0, xend = aggregate, yend = rule),
    color = onet2r_colors[["light_gray"]],
    linewidth = 1.1
  ) +
  ggplot2::geom_point(color = onet2r_colors[["teal"]], size = 4) +
  ggplot2::geom_text(
    ggplot2::aes(label = sprintf("%.3f", aggregate)),
    hjust = -0.3,
    size = 3.5,
    color = onet2r_colors[["ink"]]
  ) +
  ggplot2::scale_x_continuous(
    expand = ggplot2::expansion(mult = c(0, 0.18))
  ) +
  ggplot2::labs(
    title = "Task handling moves the headline number",
    subtitle = "Same task score, two rollup rules. The dashed line is the baseline.",
    x = "Aggregate score",
    y = NULL
  ) +
  onet2r_theme()
```

The diagnostic will not tell you which task score is right. It reports how far
the headline number moves when non-substantive plumbing changes.
