---
title: "Adding OEWS Wage and Employment Context"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding OEWS Wage and Employment Context}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(onet2r)
library(dplyr)
```

O&#42;NET tells you what occupations do: their tasks, abilities, skills, knowledge,
and work contexts. The Bureau of Labor Statistics Occupational Employment and
Wage Statistics (OEWS) program tells you how large those occupations are in the
labor market and what they pay. Joining the two lets you move from an
occupation-level descriptor table to a labor-market-weighted question.

This walkthrough asks: among occupations in a small O&#42;NET ability panel, which
ability ratings matter most once we account for national employment and wages?
The OEWS file is bundled with the package so the article runs without network
access. In production, call `onet_oews_national()` without `path` to download
the public BLS file.

```{r paths}
sample_oews <- system.file(
  "extdata",
  "oews-national-sample.csv",
  package = "onet2r"
)

archive_303 <- system.file(
  "extdata",
  "onet-mini",
  "db_30_3_text",
  package = "onet2r"
)
```

## Read OEWS Estimates

`onet_oews_national()` normalizes BLS OEWS files into snake_case columns and
parses formatted employment and wage fields into numeric values.

```{r read-oews}
oews <- onet_oews_national(year = 2024, path = sample_oews)
oews
```

Key columns for most O&#42;NET joins are `occ_code`, `tot_emp`, and `a_median`.

```{r inspect-oews}
oews |>
  select(occ_code, occ_title, tot_emp, a_median, h_median)
```

## Get O&#42;NET Occupations from an Actual Package Function

The occupation list below comes from `onet_archive_read()`, not a hand-built
example table. It reads a local archive-format O&#42;NET `Abilities` table and then
keeps one row per occupation.

```{r occupations-from-archive}
abilities <- onet_archive_read(
  "30.3",
  "Abilities",
  path = archive_303,
  release_date = "2026-05-01"
)

occupations <- abilities |>
  distinct(code = onet_soc_code, title)

occupations
```

## Join OEWS to O&#42;NET Occupation Codes

O&#42;NET-SOC codes commonly include a detail suffix such as `.00`. OEWS uses
six-digit SOC codes. `onet_join_oews()` creates a normalized `soc_code` and joins
on that value.

```{r join-oews}
occupation_wages <- onet_join_oews(occupations, oews = oews)

occupation_wages |>
  select(code, title, soc_code, tot_emp, a_median, h_median)
```

The join also tells you where wage context is missing. Missing rows usually mean
you need a different OEWS geography, a different year, or a crosswalk between
classification vintages.

```{r missing-oews}
occupation_wages |>
  transmute(
    code,
    title,
    has_oews_match = !is.na(tot_emp)
  )
```

## Compute Employment-Weighted O&#42;NET Summaries

If you average ability scores across occupations without weights, a niche
occupation counts the same as an occupation employing millions of workers.
`onet_weighted_summary()` uses OEWS employment, and optionally wage, to summarize
O&#42;NET descriptor rows at the labor-market level.

```{r weighted-summary}
weighted_abilities <- onet_weighted_summary(
  abilities,
  group = c("element_id", "element_name"),
  value = "data_value",
  occupation_code = "onet_soc_code",
  oews = oews
)

weighted_abilities
```

In this example, `n_occupations` is the number of O&#42;NET occupations with a
matching OEWS estimate for each ability row. `total_weight` is the summed OEWS
employment used in the weighted mean.

## Prioritize Occupation-Level Rows

A practical analyst often wants to know where a descriptor is both high and
attached to a large workforce. The next table joins ability rows to OEWS and
sorts by employment-weighted ability contribution.

```{r priority-table}
ability_context <- abilities |>
  onet_join_oews(oews = oews, by = "onet_soc_code") |>
  mutate(weighted_value = data_value * tot_emp) |>
  select(
    onet_soc_code,
    title,
    element_name,
    data_value,
    tot_emp,
    a_median,
    weighted_value
  ) |>
  arrange(desc(weighted_value))

head(ability_context, 10)
```

This table is not a causal estimate. It is a screening tool. It helps answer
questions such as: "Which O&#42;NET ability ratings are attached to the most
workers?" or "Which occupations should we inspect first when designing training
or workforce-transition analyses?"

## PUMS Employment Weights

If your workflow uses ACS or CPS microdata instead of OEWS employment counts,
`onet_pums_employment_weights()` converts person-level records into SOC-level
weights. The tiny input below is intentionally minimal: in a real analysis,
`SOCP` and `PWGTP` would usually come from `tidycensus::get_pums()`.

```{r pums-weights}
pums <- tibble::tibble(
  SOCP = c("151252", "151252", "291141"),
  PWGTP = c(120, 80, 200)
)

onet_pums_employment_weights(pums)
```

Use OEWS when you want official occupation employment and wage estimates. Use
PUMS-derived weights when your analysis needs custom populations, geographies,
or demographic slices that OEWS does not publish directly.
