---
title: "Adding OEWS wage and employment context"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding OEWS wage and employment context}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

O*NET describes the tasks, skills, abilities, and work contexts associated with
occupations. The Bureau of Labor Statistics Occupational Employment and Wage
Statistics (OEWS) program adds labor-market scale: employment counts and wage
estimates by SOC occupation code.

This vignette uses a small bundled OEWS sample so it can run without network
access. For real analyses, call `onet_oews_national()`, `onet_oews_state()`,
`onet_oews_metro()`, or `onet_oews_industry()` without `path`.

```{r setup}
library(onet2r)
library(dplyr)

sample_oews <- system.file(
  "extdata",
  "oews-national-sample.csv",
  package = "onet2r"
)
```

## Read OEWS estimates

```{r read-oews}
oews <- onet_oews_national(year = 2024, path = sample_oews)
oews
```

The parser normalizes column names to snake_case and converts OEWS numeric
fields such as employment and annual median wages from formatted strings to
numeric columns.

```{r inspect-oews}
str(oews)
```

## Join OEWS to O*NET occupation codes

O*NET-SOC codes commonly include a detail suffix, such as `.00`. OEWS uses
six-digit SOC codes. `onet_join_oews()` creates a `soc_code` column and joins
on that normalized code.

```{r join-oews}
occupations <- tibble::tibble(
  code = c("15-1252.00", "29-1141.00", "11-1011.00"),
  title = c("Software Developers", "Registered Nurses", "Chief Executives")
)

occupation_wages <- onet_join_oews(occupations, oews = oews)
occupation_wages
```

## Employment-weighted O*NET summaries

Many research workflows need to turn occupation-level O*NET records into
labor-market-weighted measures. For example, the same skill can appear in
multiple occupations with different skill ratings and very different employment
levels.

```{r weighted-summary}
skill_rows <- tibble::tibble(
  code = c("15-1252.00", "15-1252.00", "29-1141.00", "29-1141.00"),
  element_id = c("2.A.1.a", "2.A.1.b", "2.A.1.a", "2.A.1.b"),
  element_name = c(
    "Reading Comprehension",
    "Active Listening",
    "Reading Comprehension",
    "Active Listening"
  ),
  data_value = c(4.12, 4.00, 3.88, 4.25)
)

onet_weighted_summary(
  skill_rows,
  group = c("element_id", "element_name"),
  value = "data_value",
  oews = oews
)
```

The output includes:

- `n_records`: number of rows used in each grouped summary
- `n_occupations`: number of occupations contributing to the summary
- `total_weight`: sum of employment weights
- `weighted_mean`: employment-weighted mean of the selected O*NET value
- `wage_weighted_mean`: mean weighted by employment multiplied by annual median wage

## PUMS employment weights

If your workflow uses ACS or CPS microdata instead of OEWS employment counts,
`onet_pums_employment_weights()` converts individual records to SOC-level
weights. The `SOCP` and `PWGTP` names match common ACS PUMS output from
`tidycensus::get_pums()`.

```{r pums-weights}
pums <- tibble::tibble(
  SOCP = c("151252", "151252", "291141"),
  PWGTP = c(120, 80, 200)
)

onet_pums_employment_weights(pums)
```

For live ACS pulls, start with `tidycensus::get_pums()` and then pass the
resulting data frame to `onet_pums_employment_weights()`. The site reports
whether `tidycensus` is installed, then keeps the example local and
reproducible.

```{r tidycensus-status}
tibble::tibble(
  optional_package = "tidycensus",
  installed = requireNamespace("tidycensus", quietly = TRUE)
)
```

## Real OEWS downloads

BLS publishes annual public downloads. In a live workflow, call
`onet_oews_national()`, `onet_oews_state()`, `onet_oews_metro()`, or
`onet_oews_industry()` without `path` to download and cache the corresponding
file. This article uses the bundled sample and shows the same columns you should
inspect after a live download.

If automated downloads are rate-limited, download the ZIP manually from
the BLS OEWS public data pages and pass the local path:

```{r manual-download-shape}
head(onet_oews_national(2024, path = sample_oews), 5)
```
