---
title: "Adding OEWS Wage and Employment Context"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding OEWS Wage and Employment Context}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
helper_paths <- c(
  "inst/vignette-helpers.R",
  "vignette-helpers.R",
  system.file("vignette-helpers.R", package = "onet2r")
)
helper_path <- helper_paths[nzchar(helper_paths) & file.exists(helper_paths)][[1]]
source(helper_path)

sample_oews <- system.file(
  "extdata",
  "oews-national-sample.csv",
  package = "onet2r"
)

archive_303 <- system.file(
  "extdata",
  "onet-mini",
  "db_30_3_text",
  package = "onet2r"
)
```

O&#42;NET tells you what occupations do: their tasks, abilities, skills, knowledge,
and work contexts. The Bureau of Labor Statistics Occupational Employment and
Wage Statistics (OEWS) program tells you how large those occupations are in the
labor market and what they pay. Joining the two lets you move from an
occupation-level descriptor table to a labor-market-weighted question.

This walkthrough asks: among occupations in a small O&#42;NET ability panel, what
does a national employment-weighted ability score look like, and what changes if
we use a custom PUMS-style population instead?

## Read OEWS Estimates

`onet_oews_national()` normalizes BLS OEWS files into snake_case columns and
parses formatted employment and wage fields into numeric values.

```{r read-oews}
oews <- onet_oews_national(year = 2024, path = sample_oews)

oews |>
  select(occ_code, occ_title, tot_emp, a_median, h_median) |>
  onet_kable()
```

## Build a Reference-SOC Weight Panel

OEWS uses 6-digit SOC codes. O&#42;NET archive tables use 8-digit O&#42;NET-SOC detail
codes. The weight-panel helper records the source and reference taxonomy so the
join is auditable.

```{r oews-weight-panel}
oews_weights <- onet_weight_panel_oews(oews, year = 2024)

oews_weights |>
  onet_kable()
```

## Get O&#42;NET Occupation Scores from an Archive

The occupation scores below come from `onet_archive_read()`, not a hand-built
example table. We use Oral Comprehension as a concrete descriptor because it is
available in the bundled fixture.

```{r occupation-scores}
abilities <- onet_archive_read(
  "30.3",
  "Abilities",
  path = archive_303,
  release_date = "2026-05-01"
)

oral_scores <- abilities |>
  filter(element_id == "1.A.1.a.1") |>
  transmute(
    onet_soc_code,
    title,
    measure_score = data_value
  )

oral_scores |>
  onet_kable()
```

## Aggregate with OEWS Employment

```{r aggregate-oews}
oral_oews <- onet_measure_aggregate(
  oral_scores,
  oews_weights,
  measure_id = "oral_comprehension_fixture"
)

oral_oews |>
  select(-coverage, -provenance) |>
  onet_kable()

onet_provenance(oral_oews) |>
  onet_kable()
```

The aggregate is the employment-weighted score for occupations covered by both
the O&#42;NET fixture and the OEWS sample. Coverage tells you how much of the weight
panel was included.

```{r aggregate-coverage}
onet_coverage(oral_oews) |>
  onet_kable()
```

## Inspect Occupation Contributions

```{r contribution-table}
contributions <- oral_scores |>
  mutate(reference_soc_code = sub("\\.\\d{2}$", "", onet_soc_code)) |>
  inner_join(oews_weights, by = join_by(reference_soc_code), relationship = "many-to-one") |>
  mutate(weighted_score = measure_score * employment) |>
  arrange(desc(weighted_score)) |>
  select(title, reference_soc_code, measure_score, employment, weight_share, weighted_score)

contributions |>
  onet_kable()
```

```{r contribution-chart, fig.alt = "Horizontal bar chart showing employment-weighted ability contributions by occupation."}
ggplot2::ggplot(contributions, ggplot2::aes(
  x = weighted_score,
  y = stats::reorder(title, weighted_score)
)) +
  ggplot2::geom_col(fill = onet2r_colors[["teal"]], width = 0.65) +
  ggplot2::scale_x_continuous(
    labels = scales::label_number(scale_cut = scales::cut_short_scale()),
    expand = ggplot2::expansion(mult = c(0, 0.05))
  ) +
  ggplot2::labs(
    title = "Which Occupations Drive the Weighted Score?",
    subtitle = "Contribution is the occupation score times employment.",
    x = "Score times employment",
    y = NULL
  ) +
  onet2r_theme()
```

## Compare OEWS with a PUMS-Style Weight Panel

OEWS is the right default for official occupation employment and wage context.
PUMS-derived weights are useful when the target population is a custom sample,
geography, or demographic cell.

```{r pums-weights}
pums <- tibble::tibble(
  SOCP = c("151252", "151252", "291141", "291141", "111011"),
  PWGTP = c(120, 80, 90, 110, 20),
  sex = c("F", "M", "F", "M", "F")
)

pums_weights <- onet_weight_panel_pums(
  pums,
  year = 2022,
  group = "sex"
)

pums_weights |>
  onet_kable()
```

Because this PUMS-style panel has cells, pick one cell before aggregating.

```{r pums-aggregate}
oral_pums_f <- onet_measure_aggregate(
  oral_scores,
  pums_weights,
  measure_id = "oral_comprehension_fixture",
  cell = list(sex = "F")
)

tibble::tibble(
  weight_source = c("OEWS national", "PUMS-style F cell"),
  aggregate = c(oral_oews$aggregate, oral_pums_f$aggregate),
  coverage = c(
    oral_oews$employment_coverage_share,
    oral_pums_f$employment_coverage_share
  )
) |>
  onet_kable()
```

The point is not that one source is always better. OEWS gives official labor
market estimates. PUMS gives custom cells. The package makes the weight source,
year, taxonomy, and coverage explicit so readers can see what changed.
