---
title: "Adding OEWS Wage and Employment Context"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding OEWS Wage and Employment Context}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
helper_paths <- c(
  system.file("vignette-helpers.R", package = "onet2r"),
  "inst/vignette-helpers.R",
  "../inst/vignette-helpers.R",
  "vignette-helpers.R"
)
helper_path <- helper_paths[nzchar(helper_paths) & file.exists(helper_paths)]
if (length(helper_path) == 0) {
  stop("Could not find vignette-helpers.R.", call. = FALSE)
}
source(helper_path[[1]])

sample_oews <- onet2r_inst_path("extdata", "oews-national-sample.csv")

archive_303 <- onet2r_inst_path("extdata", "onet-mini", "db_30_3_text")
```

O&#42;NET tells you what occupations do: their tasks, abilities, skills, knowledge,
and work contexts. The Bureau of Labor Statistics Occupational Employment and
Wage Statistics (OEWS) program tells you how large those occupations are in the
labor market and what they pay. Joining the two lets you move from an
occupation-level descriptor table to a labor-market-weighted question.

This walkthrough asks: among occupations in a small O&#42;NET ability panel, what
does a national employment-weighted ability score look like, and what changes if
we use a custom PUMS-style population instead?

## Read OEWS Estimates

`onet_oews_national()` normalizes BLS OEWS files into snake_case columns and
parses formatted employment and wage fields into numeric values.

BLS sometimes rejects automated ZIP downloads even when the same OEWS URL works
in a browser. In interactive sessions, `onet_oews_national()` opens the official
BLS URL in your browser, waits for the matching ZIP to appear in your Downloads
folder, validates it, and copies it into the package cache. You can set
`options(onet2r.oews_download_dir = "path/to/downloads")` for a different folder
or pass the file explicitly with `path`, which is what this article does so it
can build without network access. Passing `path` is also the fallback if BLS
changes the OEWS URL or file name.

```{r read-oews}
oews <- onet_oews_national(year = 2023, path = sample_oews)

oews |>
  select(occ_code, occ_title, tot_emp, a_median, h_median, a_median_topcoded) |>
  knitr::kable(digits = 3, align = "l")
```

The sample includes BLS hierarchy rows and a top-coded wage. The parser keeps
the `o_group` column, parses numeric fields, and adds flag columns when a wage
is top-coded or suppressed.

## Build a Reference-SOC Weight Panel

OEWS uses 6-digit SOC codes. O&#42;NET archive tables use 8-digit O&#42;NET-SOC detail
codes. The weight-panel helper records the source and reference taxonomy so the
join is auditable.

```{r oews-weight-panel}
oews_weights <- onet_weight_panel_oews(oews, year = 2024)

oews_weights |>
  knitr::kable(digits = 3, align = "l")
```

## Get O&#42;NET Occupation Scores from an Archive

The occupation scores below come from `onet_archive_read()`, not a hand-built
example table. We use Oral Comprehension as a concrete descriptor because it is
available in the bundled fixture.

```{r occupation-scores}
abilities <- onet_archive_read(
  "30.3",
  "Abilities",
  path = archive_303,
  release_date = "2026-05-01"
)

oral_scores <- abilities |>
  filter(element_id == "1.A.1.a.1") |>
  transmute(
    onet_soc_code,
    title,
    measure_score = data_value
  )

oral_scores |>
  knitr::kable(digits = 3, align = "l")
```

## Aggregate with OEWS Employment

```{r aggregate-oews}
oral_oews <- onet_measure_aggregate(
  oral_scores,
  oews_weights,
  measure_id = "oral_comprehension_fixture"
)

oral_oews |>
  select(-coverage, -provenance) |>
  knitr::kable(digits = 3, align = "l")

onet_provenance(oral_oews) |>
  knitr::kable(digits = 3, align = "l")
```

The aggregate is the employment-weighted score for occupations covered by both
the O&#42;NET fixture and the OEWS sample. Coverage tells you how much of the weight
panel was included.

```{r aggregate-coverage}
onet_coverage(oral_oews) |>
  knitr::kable(digits = 3, align = "l")
```

## Map O&#42;NET Occupations into OEWS Combinations

Since the May 2021 estimates, OEWS publishes most detailed SOC occupations but
reports a few only as combinations. For example, `31-1120` Home Health and
Personal Care Aides holds the SOC occupations `31-1121` and `31-1122`, which
O&#42;NET rates separately. Without a bridge, those O&#42;NET occupations match no
panel row, and the combined employment stays in the denominator as unmatched.
`onet_oews_bridge()` maps each O&#42;NET occupation to the OEWS code that
publishes it. The fixture OEWS sample has no combined codes, so this example uses
stylized scores and weights.

```{r oews-combination-bridge}
# Stylized weights and scores for illustration; not OEWS or O*NET values.
combined_weights <- tibble::tibble(
  reference_soc_code = c("29-1141", "31-1120"),
  year = 2024L,
  employment = c(300, 400),
  weight_share = c(3, 4) / 7,
  source = "OEWS",
  source_taxonomy = "2018 SOC",
  reference_taxonomy = "2018 SOC"
)
stylized_scores <- tibble::tibble(
  onet_soc_code = c("29-1141.00", "31-1121.00", "31-1122.00"),
  measure_score = c(0.2, 0.4, 0.6)
)

combination_bridge <- onet_oews_bridge(stylized_scores, combined_weights)

combination_bridge |>
  knitr::kable(align = "l")
```

```{r oews-combination-aggregate}
without_bridge <- suppressMessages(onet_measure_aggregate(
  stylized_scores,
  combined_weights,
  measure_id = "stylized_score"
))
with_bridge <- onet_measure_aggregate(
  stylized_scores,
  combined_weights,
  bridge = combination_bridge,
  measure_id = "stylized_score"
)

tibble::tibble(
  run = c("No bridge", "onet_oews_bridge()"),
  aggregate = c(without_bridge$aggregate, with_bridge$aggregate),
  employment_coverage_share = c(
    without_bridge$employment_coverage_share,
    with_bridge$employment_coverage_share
  ),
  n_occupations = c(without_bridge$n_occupations, with_bridge$n_occupations)
) |>
  knitr::kable(digits = 3, align = "l")
```

Occupations inside one combination are averaged with equal weight, because OEWS
publishes no employment split among them. With O&#42;NET 31.0 task ratings and the
May 2025 national OEWS file, the bridge raises covered employment from about 92
percent to about 98 percent, and Healthcare Support from about 46 percent to 100
percent. The combinations come from the BLS occupation definitions rather than
from the rows of the panel, so the same bridge works for state, metropolitan, and
industry panels from May 2021 on, even when some of their occupations are
suppressed.

## Inspect Occupation Contributions

```{r contribution-table}
contributions <- oral_scores |>
  mutate(reference_soc_code = sub("\\.\\d{2}$", "", onet_soc_code)) |>
  summarise(
    title = paste(sort(unique(title)), collapse = "; "),
    measure_score = mean(measure_score),
    .by = "reference_soc_code"
  ) |>
  inner_join(oews_weights, by = join_by(reference_soc_code), relationship = "many-to-one") |>
  mutate(weighted_score = measure_score * employment) |>
  arrange(desc(weighted_score)) |>
  select(title, reference_soc_code, measure_score, employment, weight_share, weighted_score)

contributions |>
  knitr::kable(digits = 3, align = "l")
```

The contribution is already the `weighted_score` column: occupation score times
employment. In this fixture a separate chart would only redraw the table.

## Compare OEWS with a PUMS-Style Weight Panel

OEWS is the right default for official occupation employment and wage context.
PUMS-derived weights are useful when the target population is a custom sample,
geography, or demographic cell.

```{r pums-weights}
pums <- tibble::tibble(
  SOCP = c("151252", "151252", "291141", "291141", "111011"),
  PWGTP = c(120, 80, 90, 110, 20),
  sex = c("F", "M", "F", "M", "F")
)

pums_weights <- onet_weight_panel_pums(
  pums,
  year = 2022,
  group = "sex"
)

pums_weights |>
  knitr::kable(digits = 3, align = "l")
```

Because this PUMS-style panel has cells, pick one cell before aggregating.

For real ACS PUMS, filter to the employment universe before building weights.
The common starting point is employed civilians, often `ESR %in% c(1, 2)` and
age 16 or older. `SOCP` is the SOC field. `OCCP` is a Census occupation recode
and needs a separate crosswalk before it can be used with O&#42;NET or OEWS.

```{r pums-fetch-example, eval = FALSE}
pums <- tidycensus::get_pums(
  variables = c("SOCP", "PWGTP", "ESR", "AGEP"),
  state = "WA",
  year = 2023,
  survey = "acs1"
) |>
  filter(ESR %in% c("1", "2"), AGEP >= 16)

pums_weights <- onet_weight_panel_pums(pums, year = 2023)
```

```{r pums-aggregate}
oral_pums_f <- onet_measure_aggregate(
  oral_scores,
  pums_weights,
  measure_id = "oral_comprehension_fixture",
  cell = list(sex = "F")
)

tibble::tibble(
  weight_source = c("OEWS national", "PUMS-style F cell"),
  aggregate = c(oral_oews$aggregate, oral_pums_f$aggregate),
  coverage = c(
    oral_oews$employment_coverage_share,
    oral_pums_f$employment_coverage_share
  )
) |>
  knitr::kable(digits = 3, align = "l")
```

Neither source is always better. OEWS gives official labor market estimates;
PUMS gives custom cells. The package records the weight source, year, taxonomy,
and coverage so a reader can see what changed.
