---
title: "Validating onet2r Outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating onet2r Outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(onet2r)
library(dplyr)

sample_oews <- system.file(
  "extdata",
  "oews-national-sample.csv",
  package = "onet2r"
)

archive_303 <- system.file(
  "extdata",
  "onet-mini",
  "db_30_3_text",
  package = "onet2r"
)
```

The package test suite checks implementation details with mocked responses and
small deterministic files. For real projects, it is also useful to inspect output
shape: Are expected columns present? Are rows non-empty where they should be?
Are list endpoints returning the fields users need?

`onet2r` includes a practical validation script that maintainers can run from
the source tree with `Rscript inst/examples/validate-outputs.R`. From an
installed package, source the file returned by:

```{r script-path}
system.file("examples", "validate-outputs.R", package = "onet2r")
```

## What the Script Checks

The validation script always checks deterministic local outputs:

- reading bundled OEWS sample data;
- reading a local O&#42;NET archive-format table;
- joining O&#42;NET occupation codes to OEWS estimates;
- computing employment- and wage-weighted O&#42;NET summaries;
- configuring cache and rate-limit settings.

When `ONET_API_KEY` is available, it also performs live smoke checks across the
main exported O&#42;NET functions. Each result is inspected for expected type,
minimum row count, and important columns.

## Example: Local Validation without an API Key

The local validation below uses package functions rather than copied output.

```{r local-validation}
oews <- onet_oews_national(2024, path = sample_oews)

abilities <- onet_archive_read(
  "30.3",
  "Abilities",
  path = archive_303,
  release_date = "2026-05-01"
)

occupations <- abilities |>
  distinct(code = onet_soc_code, title)

joined <- onet_join_oews(occupations, oews = oews)

weighted <- onet_weighted_summary(
  abilities,
  group = c("element_id", "element_name"),
  value = "data_value",
  occupation_code = "onet_soc_code",
  oews = oews
)

tibble::tibble(
  check = c(
    "OEWS output is a tibble",
    "archive output has source dates",
    "join adds normalized SOC code",
    "weighted summary has employment weights"
  ),
  passed = c(
    inherits(oews, "tbl_df"),
    all(c("source_date", "data_value") %in% names(abilities)),
    all(c("soc_code", "tot_emp", "a_median") %in% names(joined)),
    all(c("total_weight", "weighted_mean") %in% names(weighted))
  )
)

head(joined, 5)
head(weighted, 5)
```

## Example: Live Validation Status

Live validation only runs when a user supplies an API key. The site still
executes this chunk and reports whether live checks are available in the current
environment.

```{r live-validation-status}
tibble::tibble(
  check = "ONET_API_KEY available for live API validation",
  passed = nzchar(Sys.getenv("ONET_API_KEY"))
)
```

The goal is not to replace unit tests. It is to give maintainers and users a
repeatable, practical way to confirm that live API outputs still match the
shapes used in examples, vignettes, and downstream analyses.
