---
title: "Validating onet2r outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating onet2r outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package test suite checks implementation details with mocked responses and
small deterministic examples. For real projects, it is also useful to inspect
output shape: Are expected columns present? Are rows non-empty where they should
be? Are list endpoints returning the fields users need?

`onet2r` includes a practical validation script that maintainers can run from
the source tree with `Rscript inst/examples/validate-outputs.R`. From an
installed package, source the file returned by:

```{r script-path}
system.file("examples", "validate-outputs.R", package = "onet2r")
```

## What the Script Checks

The validation script always checks deterministic local outputs:

- reading bundled OEWS sample data
- joining O*NET occupation codes to OEWS estimates
- computing employment- and wage-weighted O*NET summaries
- configuring cache and rate-limit settings

When `ONET_API_KEY` is available, it also performs live smoke checks across the
main exported O*NET functions. Each result is inspected for expected type,
minimum row count, and important columns.

## Example: Local Validation without an API Key

```{r local-validation}
library(onet2r)
library(tibble)

sample_oews <- system.file(
  "extdata",
  "oews-national-sample.csv",
  package = "onet2r"
)

oews <- onet_oews_national(2024, path = sample_oews)

validation <- tibble(
  check = c(
    "OEWS output is a tibble",
    "OEWS has wage and employment columns",
    "join adds normalized SOC code"
  ),
  passed = c(
    inherits(oews, "tbl_df"),
    all(c("occ_code", "tot_emp", "a_median") %in% names(oews)),
    TRUE
  )
)

occupations <- tibble(
  code = c("15-1252.00", "29-1141.00"),
  title = c("Software Developers", "Registered Nurses")
)

joined <- onet_join_oews(occupations, oews = oews)
validation$passed[3] <- all(c("soc_code", "tot_emp", "a_median") %in% names(joined))

validation
head(joined, 5)
```

## Example: Live Validation Status

Live validation only runs when a user supplies an API key. The site still
executes this chunk and reports whether live checks are available in the current
environment.

```{r live-validation-status}
tibble(
  check = "ONET_API_KEY available for live API validation",
  passed = nzchar(Sys.getenv("ONET_API_KEY"))
)
```

The goal is not to replace unit tests. It is to give maintainers and users a
repeatable, practical way to confirm that live API outputs still match the
shapes used in examples, vignettes, and downstream analyses.
