---
title: "Validating onet2r Outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating onet2r Outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(tibble.print_min = 4, tibble.print_max = 8)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
helper_paths <- c("vignettes/_helpers.R", "_helpers.R")
source(helper_paths[file.exists(helper_paths)][[1]])

sample_oews <- system.file(
  "extdata",
  "oews-national-sample.csv",
  package = "onet2r"
)

archive_303 <- system.file(
  "extdata",
  "onet-mini",
  "db_30_3_text",
  package = "onet2r"
)
```

The package test suite checks implementation details with mocked responses and
small deterministic files. For real projects, it is also useful to inspect
output shape: Are expected columns present? Are rows non-empty where they should
be? Are list endpoints returning the fields users need?

`onet2r` includes a practical validation script that maintainers can run from
the source tree with `Rscript inst/examples/validate-outputs.R`. From an
installed package, source the file returned by:

```{r script-path}
system.file("examples", "validate-outputs.R", package = "onet2r")
```

## What the Script Checks

The validation script always checks deterministic local outputs:

- reading bundled OEWS sample data;
- reading a local O&#42;NET archive-format table;
- creating a reference-SOC OEWS weight panel;
- aggregating an occupation score with coverage and provenance;
- configuring cache and rate-limit settings.

When `ONET_API_KEY` is available, it also performs live smoke checks across the
main exported O&#42;NET functions. Each result is inspected for expected type,
minimum row count, and important columns.

## Example: Local Validation without an API Key

The local validation below uses package functions rather than copied output.

```{r local-validation}
oews <- onet_oews_national(2024, path = sample_oews)
weights <- onet_weight_panel_oews(oews, year = 2024)

abilities <- onet_archive_read(
  "30.3",
  "Abilities",
  path = archive_303,
  release_date = "2026-05-01"
)

oral_scores <- abilities |>
  filter(element_id == "1.A.1.a.1") |>
  transmute(onet_soc_code, measure_score = data_value)

aggregate <- onet_measure_aggregate(
  oral_scores,
  weights,
  measure_id = "oral_comprehension_fixture"
)

tibble::tibble(
  check = c(
    "OEWS output is a tibble",
    "archive output has source dates",
    "weight panel has reference SOC codes",
    "aggregate has provenance",
    "aggregate has coverage"
  ),
  passed = c(
    inherits(oews, "tbl_df"),
    all(c("source_date", "data_value") %in% names(abilities)),
    all(c("reference_soc_code", "employment") %in% names(weights)),
    nrow(onet_provenance(aggregate)) == 1,
    nrow(onet_coverage(aggregate)) == 1
  )
) |>
  onet_kable()

weights |>
  head(5) |>
  onet_kable()

aggregate |>
  select(-coverage, -provenance) |>
  onet_kable()
```

## Example: Live Validation Status

Live validation only runs when a user supplies an API key. The site still
executes this chunk and reports whether live checks are available in the current
environment.

```{r live-validation-status}
tibble::tibble(
  check = "ONET_API_KEY available for live API validation",
  passed = nzchar(Sys.getenv("ONET_API_KEY"))
) |>
  onet_kable()
```

The goal is not to replace unit tests. It is to give maintainers and users a
repeatable, practical way to confirm that live API outputs still match the
shapes used in examples, vignettes, and downstream analyses.
