---
title: "Resurvey Cycles and Content Change"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Resurvey Cycles and Content Change}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
options(
  cli.num_colors = 1,
  crayon.enabled = FALSE,
  pillar.bold = FALSE,
  pillar.subtle = FALSE
)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
helper_paths <- c(
  system.file("vignette-helpers.R", package = "onet2r"),
  "inst/vignette-helpers.R",
  "../inst/vignette-helpers.R",
  "vignette-helpers.R"
)
helper_path <- helper_paths[nzchar(helper_paths) & file.exists(helper_paths)]
if (length(helper_path) == 0) {
  stop("Could not find vignette-helpers.R.", call. = FALSE)
}
source(helper_path[[1]])

archive_base <- onet2r_inst_path("extdata", "onet-mini")
```

O&#42;NET does not re-rate every occupation in every release. Its incumbent
survey rotates across part of the taxonomy each year, and the rest of the
database is carried forward. A task rating that did not move between two
releases may simply not have been measured again, and a comparison only carries
information about content change when the occupation was actually resurveyed.

Three verbs make that structure explicit:

- `onet_resurvey_panel()` exposes the occupation survey clock, which is the
  newest survey `source_date` in each release.
- `onet_condition_on_resurvey()` labels every row with the reason it is or is not
  at risk of showing change.
- `onet_content_change()` computes seam-aware task-set and rating metrics between
  releases.

They supply denominators and metrics. Modeling, weighting, and inference stay
with you.

## A Synthetic Three-Release Panel

The panel below is synthetic. The O&#42;NET-SOC codes and titles are real, but
the task ids, Importance ratings, and dates are invented so that each pattern
appears once: a resurvey that changes ratings, a carry-forward that looks like
stability, a resurvey that replaces a task, and a suppressed estimate.

```{r synthetic-panel}
release_rows <- function(version, date, task_id, value, source_date,
                         suppress = "N") {
  tibble::tibble(
    release_version = version,
    release_date = as.Date(date),
    soc_vintage = "2019",
    onet_soc_code = rep(c("11-3031.01", "13-2011.00", "43-3031.00"), each = 2),
    title = rep(
      c(
        "Treasurers and Controllers",
        "Accountants and Auditors",
        "Bookkeeping, Accounting, and Auditing Clerks"
      ),
      each = 2
    ),
    task_id = task_id,
    scale_id = "IM",
    data_value = value,
    source_date = as.Date(source_date),
    domain_source = "Incumbent",
    recommend_suppress = suppress
  )
}

panel <- bind_rows(
  release_rows(
    "28.0", "2023-08-01",
    task_id = c("101", "102", "201", "202", "301", "302"),
    value = c(4.2, 3.1, 4.5, 3.8, 3.9, 2.7),
    source_date = rep(c("2021-06-01", "2021-06-01", "2019-07-01"), each = 2)
  ),
  release_rows(
    "29.0", "2024-08-01",
    task_id = c("101", "102", "201", "202", "301", "302"),
    value = c(4.4, 3.4, 4.5, 3.8, 3.9, 2.7),
    source_date = rep(c("2024-03-01", "2021-06-01", "2019-07-01"), each = 2)
  ),
  release_rows(
    "30.0", "2025-08-01",
    task_id = c("101", "102", "201", "203", "301", "302"),
    value = c(4.4, 3.4, 4.6, 3.2, 3.9, 2.7),
    source_date = rep(c("2024-03-01", "2025-04-01", "2019-07-01"), each = 2),
    suppress = c("N", "N", "N", "N", "Y", "N")
  )
)

panel |>
  summarise(
    tasks = paste(task_id, collapse = ", "),
    importance = paste(data_value, collapse = ", "),
    source_date = first(source_date),
    .by = c(release_version, onet_soc_code)
  ) |>
  knitr::kable(align = "l")
```

In real work the same frame comes from `onet_panel("Task Ratings", ...)`, which
carries `source_date`, `domain_source`, and `recommend_suppress` from the archive
files.

## Read the Survey Clock

```{r survey-clock}
resurvey <- onet_resurvey_panel(panel)

resurvey |>
  distinct(
    onet_soc_code,
    release_version,
    occ_survey_date,
    prev_survey_date,
    resurvey_event,
    cycle_index,
    age_resolved
  ) |>
  arrange(onet_soc_code, release_version) |>
  knitr::kable(digits = 2, align = "l")
```

A resurvey event means the survey clock advanced since the occupation's previous
release. Treasurers and Controllers were resurveyed in 29.0, which resolved 2.75
years of staleness. Accountants and Auditors kept their 2021 ratings in 29.0 and
were resurveyed in 30.0. The clerks were never resurveyed in this window, and
the first release has no prior release to compare against.

## Build the At-Risk Set

```{r at-risk}
conditioned <- onet_condition_on_resurvey(resurvey)

conditioned |>
  count(release_version, onet_soc_code, selection_reason, at_risk) |>
  knitr::kable(align = "l")
```

Labels are assigned by precedence, so structural exclusions win over the
resurvey signal: `taxonomy_seam` first, then `suppressed`, then `resurveyed`,
then `unrevisited`. Only `resurveyed` rows are at risk. Pass
`at_risk_only = TRUE` to keep just those rows.

```{r task-status-chart, fig.height = 3.8, fig.alt = "Tile chart with one row per synthetic occupation and task and one column per release. Tiles are colored by selection reason: resurveyed, unrevisited, or suppressed."}
short_title <- c(
  "11-3031.01" = "Treasurers",
  "13-2011.00" = "Accountants",
  "43-3031.00" = "Bookkeeping clerks"
)

task_status <- conditioned |>
  mutate(
    row_label = paste0(short_title[onet_soc_code], ", task ", task_id),
    row_label = factor(row_label, levels = rev(unique(row_label)))
  )

ggplot2::ggplot(
  task_status,
  ggplot2::aes(x = release_version, y = row_label, fill = selection_reason)
) +
  ggplot2::geom_tile(color = onet2r_colors[["bg"]], linewidth = 1.2) +
  ggplot2::scale_fill_manual(
    values = c(
      resurveyed = onet2r_colors[["teal"]],
      unrevisited = onet2r_colors[["light_gray"]],
      taxonomy_seam = onet2r_colors[["amber"]],
      suppressed = onet2r_colors[["rose"]]
    ),
    name = "Selection reason"
  ) +
  ggplot2::guides(fill = ggplot2::guide_legend(title.position = "top")) +
  ggplot2::labs(
    title = "Only resurveyed rows are at risk of showing change",
    subtitle = "Synthetic panel: one tile per task and release.",
    x = "Release",
    y = NULL
  ) +
  onet2r_theme() +
  ggplot2::theme(panel.grid.major.x = ggplot2::element_blank())
```

## Measure Content Change

```{r content-change}
changes <- onet_content_change(panel)

changes |>
  select(
    onet_soc_code,
    from_release,
    to_release,
    n_added,
    n_dropped,
    n_retained,
    jaccard,
    churn_rate,
    rating_delta_l2,
    cosine,
    safely_comparable
  ) |>
  knitr::kable(digits = 3, align = "l")
```

`jaccard` is the retained share of the task union and `churn_rate` is
`1 - jaccard`. `rating_delta_l2` is the size of the rating change among retained
tasks. `cosine` compares the rating vectors over the task union with zero fill,
so it responds to both turnover and rating change.

## Keep Only Comparisons That Carry Information

Content metrics and resurvey labels answer different questions. Join them before
interpreting any change.

```{r informative-comparisons}
resurveyed <- conditioned |>
  summarise(resurveyed = any(at_risk), .by = c(onet_soc_code, release_version))

informative <- changes |>
  left_join(
    resurveyed,
    by = join_by(onet_soc_code, to_release == release_version),
    relationship = "one-to-one"
  ) |>
  select(
    onet_soc_code,
    from_release,
    to_release,
    resurveyed,
    churn_rate,
    rating_delta_l2,
    safely_comparable
  )

informative |>
  knitr::kable(digits = 3, align = "l")
```

Four of the six comparisons show no change at all, and none of those four were
resurveyed. Their zeros are carry-forwards, not evidence of stability. The two
comparisons that can speak to content change are the resurveys:

```{r at-risk-comparisons}
informative |>
  filter(resurveyed, safely_comparable) |>
  knitr::kable(digits = 3, align = "l")
```

For which occupations were actually re-rated in each official update cycle,
compare against the O&#42;NET Longitudinal Data Updates record from
`onet_data_updates()`.

## Crossing the Taxonomy Seam

O&#42;NET 25.1, released in November 2020, moved to the O&#42;NET-SOC 2019
taxonomy. Transition rows at that seam are analyst carry-forwards, not a new
survey. The bundled fixtures span that seam.

```{r seam-panel}
seam_panel <- onet_panel(
  "Task Ratings",
  versions = c("24.3", "25.1"),
  archives = c(
    `24.3` = file.path(archive_base, "db_24_3_text"),
    `25.1` = file.path(archive_base, "db_25_1_text")
  ),
  release_dates = c(`24.3` = "2020-08-01", `25.1` = "2020-11-01")
)

seam_panel |>
  onet_resurvey_panel() |>
  onet_condition_on_resurvey() |>
  select(
    release_version,
    onet_soc_code,
    domain_source,
    seam_in,
    seam_type,
    selection_reason
  ) |>
  knitr::kable(align = "l")
```

Every 25.1 row is labeled `taxonomy_seam`, including the Registered Nurses row
whose survey date moved from August to November 2020. A seam crossing is never
counted as a resurvey. Content metrics are still computed, but they are marked as
not safely comparable:

```{r seam-change}
onet_content_change(seam_panel) |>
  select(
    onet_soc_code,
    from_release,
    to_release,
    soc_vintage_from,
    soc_vintage_to,
    rating_delta_l2,
    seam_type,
    safely_comparable
  ) |>
  knitr::kable(digits = 3, align = "l")
```

Occupations are matched on `onet_soc_code`, so codes that split or merged at the
seam, such as 15-1132.00 becoming 15-1252.00 and 15-1253.00, have no pair here.
Bridge them with `onet_crosswalk_bridge()` and `onet_panel_reconcile()`, as
shown in `vignette("longitudinal-archives", package = "onet2r")`.

## Caller-Supplied Seams

The default seam table contains one package-verified seam, the November 2020
move to O&#42;NET-SOC 2019. If you have external evidence that comparisons
across another date need the same treatment for your channel, pass a `seams`
table. It replaces the default table, so keep the SOC seam row. Differences in
`soc_vintage` are always flagged regardless of this table.

```{r caller-seams}
caller_seams <- tibble::tibble(
  seam_type = c("soc_seam", "caller_seam"),
  seam_date = as.Date(c("2020-11-01", "2024-02-01"))
)

onet_content_change(panel, seams = caller_seams) |>
  select(onet_soc_code, from_release, to_release, seam_type, safely_comparable) |>
  knitr::kable(align = "l")
```

The same `seams` argument works in `onet_resurvey_panel()`, where rows entering
through a caller seam are labeled `taxonomy_seam` and leave the at-risk set. The
2024 date here is illustrative. A caller seam is a sensitivity choice that needs
its own justification; the package does not treat any other release, including
v21.0, as a default seam.
