---
title: "Stress-Testing a User-Supplied Exposure Measure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stress-Testing a User-Supplied Exposure Measure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
helper_paths <- c(
  "inst/vignette-helpers.R",
  "vignette-helpers.R",
  system.file("vignette-helpers.R", package = "onet2r")
)
helper_path <- helper_paths[nzchar(helper_paths) & file.exists(helper_paths)][[1]]
source(helper_path)

archive_path <- function(version) {
  system.file(
    "extdata",
    "onet-mini",
    paste0("db_", gsub(".", "_", version, fixed = TRUE), "_text"),
    package = "onet2r"
  )
}
oews_path <- system.file("extdata", "oews-national-sample.csv", package = "onet2r")
```

This article uses a stylized task score to show the sensitivity workflow. The
score is not a package-endorsed exposure measure. The package's job is to show
how much the headline number moves when non-substantive choices change.

## Build the Measure from Task Files

```{r build-measure}
tasks_243 <- onet_archive_read(
  "24.3",
  "Task Statements",
  path = archive_path("24.3"),
  release_date = "2020-08-01"
)
ratings_243 <- onet_archive_read(
  "24.3",
  "Task Ratings",
  path = archive_path("24.3"),
  release_date = "2020-08-01"
)
tasks_251 <- onet_archive_read(
  "25.1",
  "Task Statements",
  path = archive_path("25.1"),
  release_date = "2020-11-01"
)
ratings_251 <- onet_archive_read(
  "25.1",
  "Task Ratings",
  path = archive_path("25.1"),
  release_date = "2020-11-01"
)

task_scores <- tibble::tibble(
  task_id = c("1101", "1201", "1202", "2101"),
  score = c(0.75, 0.80, 0.55, 0.25)
)

measure <- onet_measure(
  task_scores,
  key = "task_id",
  score = "score",
  key_type = "task",
  universe = unique(c(tasks_243$task_id, tasks_251$task_id)),
  measure_id = "stylized_exposure",
  release_version = "multi_release_fixture"
)

onet_coverage(measure) |>
  onet_kable()
```

## Create Alternative Weight Panels

```{r weight-panels}
oews_weights <- onet_weight_panel_oews(
  onet_oews_national(2024, path = oews_path),
  year = 2024
)

pums <- tibble::tibble(
  SOCP = c("151252", "151253", "291141", "291141"),
  PWGTP = c(80, 120, 200, 80)
)
pums_weights <- onet_weight_panel_pums(pums, year = 2022)

oews_weights |>
  onet_kable()
pums_weights |>
  onet_kable()
```

## Run the Sensitivity Grid

```{r sensitivity-grid}
bridge_2010_2019 <- tibble::tibble(
  from_vintage = "mixed_fixture",
  to_vintage = "2018 SOC",
  from_onet_soc_code = c(
    "15-1132.00",
    "15-1132.00",
    "15-1252.00",
    "15-1253.00",
    "29-1141.00"
  ),
  to_onet_soc_code = c(
    "15-1252.00",
    "15-1253.00",
    "15-1252.00",
    "15-1253.00",
    "29-1141.00"
  ),
  map_type = c("split", "split", "one_to_one", "one_to_one", "one_to_one"),
  crosswalk_weight = c(0.5, 0.5, 1, 1, 1)
)

sensitivity <- onet_measure_sensitivity(
  measure,
  weight_panels = list(oews = oews_weights, pums = pums_weights),
  bridges = list(reference_bridge = bridge_2010_2019),
  task_ratings = list(`2010-vintage 24.3` = ratings_243, `2019-vintage 25.1` = ratings_251),
  task_metadata = list(`2010-vintage 24.3` = tasks_243, `2019-vintage 25.1` = tasks_251),
  include_supplemental = c(FALSE, TRUE)
)

sensitivity |>
  select(
    scenario,
    task_release,
    soc_vintage,
    weight_panel,
    include_supplemental,
    aggregate,
    employment_coverage_share,
    movement,
    movement_percent
  ) |>
  onet_kable()
```

```{r sensitivity-chart, fig.alt = "Horizontal dot chart of stylized exposure aggregates across task release, weight source, and task handling choices."}
plot_sensitivity <- sensitivity |>
  mutate(
    task_rule = if_else(include_supplemental, "core + supplemental", "core only"),
    plot_label = paste0(task_release, "  |  ", weight_panel, "  |  ", task_rule)
  )

ggplot2::ggplot(plot_sensitivity, ggplot2::aes(
  x = aggregate,
  y = stats::reorder(plot_label, aggregate),
  color = weight_panel,
  shape = task_rule
)) +
  ggplot2::geom_vline(
    xintercept = plot_sensitivity$baseline_aggregate[[1]],
    color = onet2r_colors[["slate"]],
    linetype = "dashed"
  ) +
  ggplot2::geom_point(size = 3.6) +
  ggplot2::scale_color_manual(
    values = c(oews = onet2r_colors[["teal"]], pums = onet2r_colors[["amber"]]),
    name = "Weight source"
  ) +
  ggplot2::scale_shape_manual(
    values = c("core only" = 16, "core + supplemental" = 17),
    name = "Task rule"
  ) +
  ggplot2::scale_x_continuous(
    expand = ggplot2::expansion(mult = c(0.08, 0.12))
  ) +
  ggplot2::labs(
    title = "Sensitivity to task release, weights, and task handling",
    subtitle = "Dashed line marks the first scenario in the grid.",
    x = "Aggregate score",
    y = NULL
  ) +
  onet2r_theme() +
  ggplot2::theme(legend.box = "vertical", legend.margin = ggplot2::margin())
```

## Inspect Provenance

```{r provenance}
onet_provenance(sensitivity) |>
  select(any_of(c("weight_source", "weight_year", "bridge", "measure_id"))) |>
  head(8) |>
  onet_kable()
```

If the sign, rank, or interpretation of a result depends on one plumbing choice,
say so in the write-up. Running the grid does not make a result more trustworthy;
it just shows where that result is fragile.
