---
title: "Stress-Testing a User-Supplied Exposure Measure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stress-Testing a User-Supplied Exposure Measure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)

archive_dir <- system.file("extdata", "onet-mini", "db_30_3_text", package = "onet2r")
oews_path <- system.file("extdata", "oews-national-sample.csv", package = "onet2r")
```

This article uses a stylized task score to show the sensitivity workflow. The
score is not a package-endorsed exposure measure. The package's job is to show
how much the headline number moves when non-substantive choices change.

## Build the Measure from Task Files

```{r build-measure}
tasks <- onet_archive_read(
  "30.3",
  "Task Statements",
  path = archive_dir,
  release_date = "2026-05-01"
)
ratings <- onet_archive_read(
  "30.3",
  "Task Ratings",
  path = archive_dir,
  release_date = "2026-05-01"
)

task_scores <- tibble::tibble(
  task_id = c("1001", "1002", "2001"),
  score = c(0.80, 0.40, 0.20)
)

measure <- onet_measure(
  task_scores,
  key = "task_id",
  score = "score",
  key_type = "task",
  universe = tasks$task_id,
  measure_id = "stylized_exposure",
  release_version = "30.3"
)

onet_coverage(measure)
```

## Create Alternative Weight Panels

```{r weight-panels}
oews_weights <- onet_weight_panel_oews(
  onet_oews_national(2024, path = oews_path),
  year = 2024
)

pums <- tibble::tibble(
  SOCP = c("151252", "151252", "291141", "291141"),
  PWGTP = c(80, 120, 200, 80)
)
pums_weights <- onet_weight_panel_pums(pums, year = 2022)

oews_weights
pums_weights
```

## Run the Sensitivity Grid

```{r sensitivity-grid}
sensitivity <- onet_measure_sensitivity(
  measure,
  weight_panels = list(oews = oews_weights, pums = pums_weights),
  task_ratings = ratings,
  task_metadata = tasks,
  include_supplemental = c(FALSE, TRUE)
)

sensitivity |>
  select(
    scenario,
    aggregate,
    employment_coverage_share,
    movement,
    movement_percent
  ) |>
  print(width = Inf)
```

```{r sensitivity-chart, fig.alt = "Bar chart of stylized exposure aggregates across task and weight choices."}
barplot(
  height = setNames(sensitivity$aggregate, sensitivity$scenario),
  col = c("#0f766e", "#14b8a6", "#64748b", "#94a3b8"),
  border = NA,
  las = 2,
  ylab = "Aggregate score",
  main = "Sensitivity to Weight and Task Choices"
)
```

## Inspect Provenance

```{r provenance}
onet_provenance(sensitivity)
```

If the sign, rank, or interpretation of a result depends on one plumbing choice,
that belongs in the write-up. The function does not make the result more true.
It makes the fragility visible.
