---
title: "Within-Versus-Between Decomposition"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Within-Versus-Between Decomposition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(tibble.print_min = 4, tibble.print_max = 8)
options(
  cli.num_colors = 1,
  crayon.enabled = FALSE,
  pillar.bold = FALSE,
  pillar.subtle = FALSE
)
```

```{r setup, include = FALSE}
library(onet2r)
library(dplyr)
```

Aggregate changes can come from two places. Occupations can change internally,
or employment can shift toward occupations that already had different scores.
With O&#42;NET, the within term needs an extra gate: only rows that survive the
comparability checks should be counted as safely comparable within-occupation
change.

## A Two-Occupation Example

```{r inputs}
from_scores <- tibble::tibble(
  reference_soc_code = c("15-1252", "29-1141"),
  measure_score = c(1.0, 2.0),
  safely_comparable = c(TRUE, FALSE)
)
to_scores <- tibble::tibble(
  reference_soc_code = c("15-1252", "29-1141"),
  measure_score = c(2.0, 2.5),
  safely_comparable = c(TRUE, FALSE)
)
from_weights <- tibble::tibble(
  reference_soc_code = c("15-1252", "29-1141"),
  employment = c(100, 100)
)
to_weights <- tibble::tibble(
  reference_soc_code = c("15-1252", "29-1141"),
  employment = c(150, 50)
)

from_scores
to_scores
```

The first occupation is safe for within-change attribution. The second is not,
so its within movement goes to the unclassifiable bucket.

```{r decomposition}
decomp <- onet_decompose_change(
  from_scores,
  to_scores,
  from_weights,
  to_weights
)

decomp |>
  select(component, value) |>
  print(width = Inf)

onet_coverage(decomp)
```

```{r decomposition-chart, fig.alt = "Bar chart showing within, between, interaction, and unclassifiable decomposition components."}
components <- decomp |>
  filter(component != "total_change")

barplot(
  height = setNames(components$value, components$component),
  col = c("#0f766e", "#14b8a6", "#99f6e4", "#64748b"),
  border = NA,
  las = 2,
  ylab = "Contribution to change",
  main = "Where Did the Aggregate Change Come From?"
)
abline(h = 0, col = "#334155")
```

The component rows other than `total_change` sum to the total change.

```{r sum-check}
decomp |>
  summarise(
    component_sum = sum(value[component != "total_change"]),
    total_change = value[component == "total_change"],
    difference = component_sum - total_change
  )
```

## A Demographic Split

The same decomposition can be run after users prepare weight panels for
demographic cells. The package does not download raw PUMS data in examples, but
the fixture below follows the same reference-SOC, cell, employment shape.

```{r demographic-weights}
from_weights_by_sex <- tibble::tibble(
  reference_soc_code = c("15-1252", "29-1141", "15-1252", "29-1141"),
  sex = c("F", "F", "M", "M"),
  employment = c(40, 70, 60, 30)
)
to_weights_by_sex <- tibble::tibble(
  reference_soc_code = c("15-1252", "29-1141", "15-1252", "29-1141"),
  sex = c("F", "F", "M", "M"),
  employment = c(70, 50, 80, 0)
)

split_results <- lapply(unique(from_weights_by_sex$sex), function(cell) {
  result <- onet_decompose_change(
    from_scores,
    to_scores,
    from_weights_by_sex |> filter(sex == cell),
    to_weights_by_sex |> filter(sex == cell)
  )
  result$sex <- cell
  result
}) |>
  purrr::list_rbind()

split_results |>
  select(sex, component, value) |>
  print(n = Inf, width = Inf)
```

Each cell has its own weight shift and its own unclassifiable bucket. That is
the part reviewers need to see when an aggregate result depends on vintage
bridges, source dates, or task-handling choices.
